---
title: "TermineR Data Preparation"
author: "Your Name"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: false
    embed-resources: true
    df-print: paged
    max-width: "1200px"
editor: source
execute:
  eval: true
  echo: false
  error: false
  warning: true
  message: true
  cache: true
---


# Setup and Parameters

```{r setup}
#| cache: false
# Load required libraries
suppressPackageStartupMessages({
  library(TermineR)
  library(tidyverse)
  library(here)
  library(arrow)
  library(SummarizedExperiment)
  library(ggpubr)
  library(pheatmap)
  library(RColorBrewer)
  library(mixOmics)
  library(naniar)
  library(kableExtra)
  library(janitor)
  library(sva)
})


# Set options
options(max.print = 1000, scipen = 999, stringsAsFactors = FALSE)

# Source helper functions
source(here("scr/modif_diann_adapter.R"))
source(here("scr/helper_functions.R"))

# Load shared pipeline configuration
config <- load_stage_parameters(
  stage = "data_preparation",
  config_path = default_pipeline_config_path()
)

# Create results directories
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(rds_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r define_parameters}
#| cache: false
# Parameters are loaded from config.yaml in the setup chunk.
invisible(config)
```

# Data Loading and Annotation

```{r load_data}
#| cache: true
#| message: true
# Load data based on search engine
rds_path_data <- file.path(rds_dir, paste0(pre_fix, "_", search_engine, "_data.rds"))

if (file.exists(rds_path_data)) {

  message("Loading cached ", search_engine, " data...")
  df_from_search <- read_rds(rds_path_data)
  
} else {
  message("Loading ", search_engine, " data...")
  
  # Load data based on search engine type
  df_from_search <- switch(search_engine,
    "diann" = {
      # DIA-NN data loading
      file_ext <- tools::file_ext(diann_report_location)
      
      if (file_ext == "parquet") {
        message("Using parquet adapter for DIA-NN report.parquet file...")

        diann_adapter_parquet(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )
      } else if (file_ext == "tsv") {
        message("Using standard diann_adapter for DIA-NN report.tsv file...")
        result <- diann_adapter(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )

        # Clean column names for TSV files
        result %>% rename_with(~ str_remove(., "_.*$"), starts_with(instrument))

      } else {
        stop("Unsupported DIA-NN file format. Please use either .parquet or .tsv files.")
      }
    },
    
    "fragpipe_tmt" = {
      message("Using FragPipe TMT adapter...")
      fragpipe_adapter(
        parent_dir = fragpipe_parent_dir,
        ref_sample = ref_sample,
        grouping_var = "nterm_modif_peptide",
        min_purity = min_purity,
        tmt_delta = tmt_delta
      )
    },
    
    "fragpipe_lf" = {
      message("Using FragPipe Label-Free adapter...")
      fragpipe_lf_adapter(
        parent_dir = fragpipe_lf_parent_dir,
        annotation_file_path = fragpipe_lf_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "fragpipe_heavy_light" = {
      message("Using FragPipe Heavy-Light adapter...")
      fragpipe_dda_heavy_light_adapter(
        combined_modified_peptide_file = fragpipe_hl_file,
        annotation_file_path = fragpipe_hl_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "spectronaut" = {
      message("Using Spectronaut adapter...")
      spectronaut_adapter(
        path_to_file = spectronaut_report,
        proteotypic = proteotypic_only
      )
    },
    
    stop("Unsupported search engine: ", search_engine, 
         ". Supported options: diann, fragpipe_tmt, fragpipe_lf, fragpipe_heavy_light, spectronaut")
  )
  
  write_rds(df_from_search, rds_path_data)

}

# Load experimental design (adapt based on search engine
# For other search engines, load external annotation file
experimental_design <- load_pipeline_experimental_design(
  location_annotation = location_annotation,
  search_data_colnames = colnames(df_from_search),
  instrument = instrument
)

# Load protein to gene mapping with caching
rds_path_prot2gene <- file.path(rds_dir, paste0(pre_fix, "_", search_engine, "_prot2gene.rds"))

if (file.exists(rds_path_prot2gene)) {

  message("Loading cached protein-to-gene mapping...")
  prot2gene <- read_rds(rds_path_prot2gene)

} else {

  prot2gene <- read_tsv(
    uniprot_annotation
  ) %>%
  janitor::clean_names() %>%
  dplyr::select(
    protein = entry,
    gene = gene_names_primary) %>%
  filter(
    protein %in% unique(df_from_search$protein)
  )

  write_rds(prot2gene, rds_path_prot2gene)
}
```

```{r annotate_peptides}
#| cache: true
# Annotate neo-termini
rds_path_annotated <- file.path(rds_dir, paste0(pre_fix, "_annotated_data.rds"))

if (file.exists(rds_path_annotated)) {

  message("Loading cached annotated data...")
  annotated_df_quant <- read_rds(rds_path_annotated)
  # Ensure flags exist even when loading from cache
  #annotated_df_quant <- add_interest_flags(annotated_df_quant)

} else {
  message("Annotating neo-termini...")

  annotated_df_quant <- annotate_neo_termini(
    peptides_df = df_from_search,
    fasta_location = fasta_location,
    sense = sense_protease,
    specificity = specificity_protease,
    organism = organism_annotation
  ) %>% 
    mutate(cleav_len = nchar(cleavage_sequence)) %>% 
    relocate(cleav_len, .after = processing_annotation_end) %>%
    filter(!is.na(peptide_start))

  annotated_df_quant <- annotated_df_quant %>%
    mutate(
      is_in_scope_for_interest = case_when(
        specificity %in% interesting_specificity &
          nterm_modif %in% interesting_nterm_modif ~ TRUE,
        TRUE ~ FALSE
      )
    )

  
  write_rds(annotated_df_quant, rds_path_annotated)
}
```

# Quick look at experimental annotation and samples

```{r}
experimental_design
```

```{r}
gtsummary::tbl_summary(
  experimental_design,
  by = condition)
```

# Correction of experimental metadata annonation

Some MS runs had to be excluded during the search because they seemed to be corrupted. The search crashed with those. This was tested after re-transferring the files and the search was tested in both Windows and Linux environments. Yielded the same results. We will exclude those from the experimental design here, to it matches the final list of samples in the search results.

If a patient has less than 2 replicates, we will exclude that patient entirely from the analysis.

```{r}
experimental_design_all_samples <- experimental_design

experimental_design <- filter_pipeline_experimental_design(
  experimental_design = experimental_design,
  observed_samples = colnames(annotated_df_quant),
  replicate_column = "replicate"
)
```

Keep only samples in the annotated_df_quant that are in the updated experimental_design

```{r}
samples_in_design <- experimental_design$sample

samples_searched <- colnames(annotated_df_quant) %>%
  .[str_starts(., instrument)]

# which samples should be excluded?
samples_to_exclude <- setdiff(samples_searched, samples_in_design)

# Preserve the full annotated table so we can audit features found only in library runs.
annotated_df_quant_all_samples <- annotated_df_quant

annotated_df_quant <- annotated_df_quant %>%
  dplyr::select(-all_of(samples_to_exclude))
```

# Identification Statistics

```{r identification_stats}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: true

# Calculate identification statistics
create_identification_plots <- function(data, experimental_design) {

  # Number of identifications per sample
  ids_per_sample <- data %>%
    dplyr::select(starts_with(instrument)) %>%
    pivot_longer(cols = starts_with(instrument), names_to = "sample", values_to = "value") %>%
    group_by(sample) %>%
    summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
    left_join(experimental_design, by = "sample")
  
  # Plot identifications by condition
  p1 <- ids_per_sample %>%
    ggplot(aes(x = sample, y = n_identifications, fill = condition)) +
    geom_col(position = "dodge") +
    labs(title = "Number of Identifications by Condition",
         x = "Condition", y = "Number of Identifications") +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90, hjust = 1),
      )
  
  return(list(plot = p1, data = ids_per_sample))
}

id_results <- create_identification_plots(annotated_df_quant, experimental_design)
print(id_results$plot)
```

# Median Identifications by Processing Type

Performed the search only N-term semispecific to make it faster and because of our focus on N-terminal acetylation.

```{r}
annotated_df_quant$nterm_modif %>% table()
```

```{r median_ids_processing_type}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: true

# Calculate identifications by specificity and refined processing type (excluding specific peptides)
ids_by_specificity_refined <- annotated_df_quant %>%
  filter(specificity %in% c("semi_Cterm", "semi_Nterm")) %>%
  dplyr::select(specificity, processing_type, nterm_modif, starts_with(instrument)) %>%
  pivot_longer(cols = starts_with(instrument),
               names_to = "sample",
               values_to = "value") %>%
  group_by(specificity, nterm_modif, processing_type, sample) %>%
  summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
  left_join(
    experimental_design, 
    by = "sample")

# Calculate median identifications by refined processing type
median_ids_refined_processing <- ids_by_specificity_refined %>%
  group_by(specificity, nterm_modif, processing_type, condition) %>%
  summarise(
    median_ids = median(n_identifications),
    mean_ids = mean(n_identifications),
    sd_ids = sd(n_identifications),
    n_samples = n(),
    .groups = 'drop'
  )

# Plot median identifications with facet_grid (horizontal bars)
plot_median_refined <- median_ids_refined_processing %>%
  ggplot(aes(x = median_ids, 
             y = condition,
             fill = condition)) +
  geom_col() +
  geom_errorbar(aes(xmin = median_ids - sd_ids, 
                    xmax = median_ids + sd_ids),
                width = 0.3, alpha = 0.7) +
  geom_text(aes(label = round(median_ids, 0)), 
            hjust = -0.01, vjust = -0.2, size = 2.5) +
  facet_grid(
    processing_type ~ nterm_modif + specificity, 
    scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Number of Identifications by Refined Processing Type\nBefore Sparcity Reduction and Imputation",
    subtitle = "Semi-specific peptides only, grouped by specificity and processing type",
    x = "Median number of identifications",
    y = "Condition",
    fill = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 9)
  )

print(plot_median_refined)
```

# Missing Values Analysis

```{r missing_values}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Analyze missing value patterns
missing_analysis <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  vis_miss(warn_large_data = FALSE) +
  labs(title = "Missing Value Patterns")

# Calculate missing value statistics
missing_stats <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "missing_count") %>%
  left_join(experimental_design, by = "sample")

missing_plot <- missing_stats %>%
  ggplot(aes(x = sample, y = missing_count, fill = condition)) +
  geom_col() +
  labs(
    title = "Missing Values by Sample", 
    x = "Sample", 
    y = "Missing Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(missing_plot)
```

```{r missing_values_all_samples_proportional}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

missing_stats_all_samples <- annotated_df_quant_all_samples %>%
  dplyr::select(starts_with(instrument)) %>%
  summarise(
    dplyr::across(
      everything(),
      ~ sum(is.na(.)),
      .names = "{.col}"
    )
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "sample",
    values_to = "missing_count"
  ) %>%
  dplyr::mutate(
    total_features = nrow(annotated_df_quant_all_samples),
    observed_count = total_features - missing_count,
    missing_fraction = missing_count / total_features,
    observed_fraction = observed_count / total_features,
    sample_status = dplyr::case_when(
      sample %in% experimental_design$sample ~ "Retained for downstream analysis",
      TRUE ~ "Excluded later from downstream analysis"
    )
  ) %>%
  dplyr::left_join(
    experimental_design_all_samples %>% dplyr::select(sample, condition),
    by = "sample"
  )

missing_stats_all_samples_long <- missing_stats_all_samples %>%
  dplyr::select(sample, condition, sample_status, missing_fraction, observed_fraction) %>%
  tidyr::pivot_longer(
    cols = c(missing_fraction, observed_fraction),
    names_to = "value_type",
    values_to = "fraction"
  ) %>%
  dplyr::mutate(
    value_type = dplyr::recode(
      value_type,
      missing_fraction = "Missing",
      observed_fraction = "Observed"
    ),
    value_type = factor(value_type, levels = c("Observed", "Missing")),
    sample = factor(sample, levels = missing_stats_all_samples$sample)
  )

missing_plot_all_samples_prop <- missing_stats_all_samples_long %>%
  ggplot(aes(x = sample, y = fraction, fill = value_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  scale_fill_manual(values = c("Observed" = "#47A8BD", "Missing" = "#D1495B")) +
  labs(
    title = "Proportion of Missing Values by Sample Including Excluded Samples",
    subtitle = "Samples excluded later in the workflow are retained here to show apparent sparsity before filtering",
    x = "Sample",
    y = "Share of features",
    fill = "Value status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

if (dplyr::n_distinct(missing_stats_all_samples_long$sample_status) > 1) {
  missing_plot_all_samples_prop <- missing_plot_all_samples_prop +
    facet_grid(~sample_status, scales = "free_x", space = "free_x")
}

print(missing_plot_all_samples_prop)
```

# Imputation and Data Preparation

```{r imputation}
#| cache: false
#| message: true
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height 

# Prepare data for downstream analysis
rds_path_prepared <- file.path(rds_dir, paste0(pre_fix, "_prepared_n_imputed_data.rds"))
imputation_list_path <- file.path(rds_dir, paste0(pre_fix, "_imputation_list.rds"))

if (file.exists(rds_path_prepared)) {

  message("Loading cached prepared data...")
  se_pept_imp_raw <- read_rds(rds_path_prepared)

} else {
  message("Preparing data...")
  
  # Create quantification matrix
  quant_matrix <- annotated_df_quant %>%
    dplyr::select(nterm_modif_peptide, starts_with(instrument)) %>%
    column_to_rownames("nterm_modif_peptide") %>%
    # make sure that the order of the columns in quant_matrix matches the order in sample experimental_design
    dplyr::select(all_of(experimental_design$sample)) %>%
    as.matrix()
  
  # For TMT data, work only with features present in all samples
  if (search_engine == "fragpipe_tmt") {

    message("TMT data detected: filtering for complete cases only...")
    complete_features <- complete.cases(quant_matrix)
    quant_matrix <- quant_matrix[complete_features, ]
    
    # Filter annotation data to match
    annotation_data <- annotated_df_quant %>%
      filter(nterm_modif_peptide %in% rownames(quant_matrix)) %>%
      # force the order of the rows to match the quant_matrix
      dplyr::arrange(match(nterm_modif_peptide, rownames(quant_matrix))) %>%
      dplyr::select(-starts_with(instrument))
    
    message("Retained ", nrow(quant_matrix), " complete features out of ", 
            nrow(annotated_df_quant), " total features")

    # Create SummarizedExperiment object
    se_pept_imp_raw <- SummarizedExperiment(
      assays = list(counts = quant_matrix),
      colData = experimental_design,
      rowData = annotation_data
    )

  } else {

    # Prepare annotation data for non-TMT workflows
    annotation_data <- annotated_df_quant %>%
      dplyr::select(-starts_with(instrument))

    # Create SummarizedExperiment object
    se_pept_raw <- SummarizedExperiment(
      assays = list(counts = quant_matrix),
      colData = experimental_design,
      rowData = annotation_data
    )

    # For other data types, perform imputation as before
    message("Non-TMT data: performing imputation...")

    se_pept_imp_raw_imputation_list <- terminer_imputation(
      se = se_pept_raw,
      min_fraction_condition = missing_accepted,
      tune_sigma = tune_sigma,
      tune_quantile = tune_quantile
    )

    se_pept_imp_raw <- se_pept_imp_raw_imputation_list$se

    # get summary table of missingness and imputation approach by feature observation
    imputation_summary_table <- se_pept_imp_raw_imputation_list$imputation_summary_table

    write_tsv(imputation_summary_table, 
          file.path(results_dir, paste0(pre_fix, "_imputation_summary_table.tsv")))

    # Visualization: reconstruct long imputed dataframe and plot summaries (conditional)
    if (exists("imputation_summary_table") && is.data.frame(imputation_summary_table) && nrow(imputation_summary_table) > 0) {
      message("Creating imputation diagnostic plots...")

      # Reconstruct long imputed dataframe from the imputed SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(
          cols = -nterm_modif_peptide,
           names_to = "sample", 
           values_to = "Abundance") %>%
        dplyr::left_join(
          imputation_summary_table, 
          by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      # Only plot if imputation method column present
      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        message("Creating imputation diagnostic plots...")
        try({
          # Histogram by condition and imputation method
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          # Density plot by condition and imputation method
          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          # Barplot for counts by imputation method
          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)

        # Mark that imputation plots were generated in this session
        imputation_plots_generated <- TRUE

      } else {
        message("Column 'imputation_method' not found in imputation summary; skipping imputation-method visualizations.")
      }

    } else {
      message("No imputation summary available — skipping imputation visualizations.")
    }

  }  # end non-TMT branch
  
  write_rds(se_pept_imp_raw, rds_path_prepared)
  if (exists("se_pept_imp_raw_imputation_list") && is.list(se_pept_imp_raw_imputation_list)) {
    write_rds(se_pept_imp_raw_imputation_list, imputation_list_path)
  }
}
```

```{r}
assay(se_pept_imp_raw)[1:10, 1:10]
```

```{r imputation_plots_from_cache}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# If we loaded cached prepared data earlier, imputation plots may not have been created.
# Recreate them from the saved TSV if needed.
if (!exists("imputation_plots_generated") || !isTRUE(imputation_plots_generated)) {

  if (!isTRUE(run_imputation_diagnostics)) {
    message("Imputation diagnostics disabled (run_imputation_diagnostics = FALSE). Skipping cached imputation plots.")
  } else {

  imp_sum_path <- file.path(results_dir, paste0(pre_fix, "_imputation_summary_table.tsv"))
  
  if (file.exists(imp_sum_path)) {

    tsv_mb <- file.info(imp_sum_path)$size / (1024^2)
    if (is.finite(tsv_mb) && tsv_mb > imputation_tsv_max_mb) {
      message(
        "Imputation summary TSV is large (~", round(tsv_mb, 1), " MB). ",
        "Skipping cached imputation plots. Increase imputation_tsv_max_mb or set run_imputation_diagnostics = TRUE to force."
      )
    } else {

    imputation_summary_table_cached <- readr::read_tsv(imp_sum_path, show_col_types = FALSE)
    if (is.data.frame(imputation_summary_table_cached) && nrow(imputation_summary_table_cached) > 0) {
      # Reconstruct long imputed dataframe from the (possibly cached) SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(cols = -nterm_modif_peptide, names_to = "sample", values_to = "Abundance") %>%
        dplyr::left_join(imputation_summary_table_cached, by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        try({
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)
      }
    }
    }
  }
  }
}
```

```{r sparse_feature_exclusion_plot}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

if (search_engine == "fragpipe_tmt") {
  message("Sparse-feature exclusion plot is not available for fragpipe_tmt because this workflow uses complete-case filtering instead of sparse-feature filtering.")
} else {

  imputation_list_path <- file.path(rds_dir, paste0(pre_fix, "_imputation_list.rds"))

  if ((!exists("se_pept_imp_raw_imputation_list") || !is.list(se_pept_imp_raw_imputation_list)) && file.exists(imputation_list_path)) {
    se_pept_imp_raw_imputation_list <- read_rds(imputation_list_path)
  }

  if (!exists("se_pept_imp_raw_imputation_list") || !is.list(se_pept_imp_raw_imputation_list)) {
    message("Sparse-feature metadata not available. Skipping sparse-feature exclusion plot.")
  } else {

  sparse_features <- se_pept_imp_raw_imputation_list$sparse_features %||% character(0)

  sparse_feature_exclusion_by_sample <- annotated_df_quant %>%
    dplyr::select(nterm_modif_peptide, all_of(experimental_design$sample)) %>%
    dplyr::mutate(
      sparse_filter_status = dplyr::case_when(
        nterm_modif_peptide %in% sparse_features ~ "Excluded by sparse-feature filter",
        TRUE ~ "Retained after sparse-feature filter"
      )
    ) %>%
    tidyr::pivot_longer(
      cols = -c(nterm_modif_peptide, sparse_filter_status),
      names_to = "sample",
      values_to = "Abundance"
    ) %>%
    dplyr::mutate(is_identified = !is.na(Abundance)) %>%
    dplyr::filter(is_identified) %>%
    dplyr::count(sample, sparse_filter_status, name = "n_features") %>%
    tidyr::complete(
      sample = experimental_design$sample,
      sparse_filter_status = c(
        "Retained after sparse-feature filter",
        "Excluded by sparse-feature filter"
      ),
      fill = list(n_features = 0)
    ) %>%
    dplyr::left_join(
      experimental_design %>% dplyr::select(sample, condition),
      by = "sample"
    ) %>%
    dplyr::mutate(
      sample = factor(sample, levels = experimental_design$sample),
      sparse_filter_status = factor(
        sparse_filter_status,
        levels = c(
          "Retained after sparse-feature filter",
          "Excluded by sparse-feature filter"
        )
      )
    )

  sparse_feature_exclusion_plot <- sparse_feature_exclusion_by_sample %>%
    ggplot(aes(x = sample, y = n_features, fill = sparse_filter_status)) +
    geom_col() +
    labs(
      title = "Identifications Retained vs Excluded After Sparse-Feature Filtering",
      subtitle = "Each bar sums to the number of pre-filter identifications per sample",
      x = "Sample",
      y = "Number of identified features",
      fill = "Feature status"
    ) +
    scale_fill_manual(
      values = c(
        "Retained after sparse-feature filter" = "#47A8BD",
        "Excluded by sparse-feature filter" = "#D1495B"
      )
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "bottom"
    )

  if ("condition" %in% colnames(sparse_feature_exclusion_by_sample)) {
    sparse_feature_exclusion_plot <- sparse_feature_exclusion_plot +
      facet_grid(~condition, scales = "free_x", space = "free_x")
  }

  print(sparse_feature_exclusion_plot)
  }
}
```
