---
title: "TermineR Exploratory Analysis"
author: "Your Name"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: false
    embed-resources: true
    df-print: paged
    max-width: "1200px"
editor: source
execute:
  eval: true
  echo: false
  error: false
  warning: false
  message: false
  cache: true
---

# Setup and Parameters

```{r setup}
#| cache: false
# Load required libraries
suppressPackageStartupMessages({
  library(TermineR)
  library(tidyverse)
  library(here)
  library(arrow)
  library(SummarizedExperiment)
  library(ggpubr)
  library(pheatmap)
  library(RColorBrewer)
  library(mixOmics)
  library(naniar)
  library(kableExtra)
  library(janitor)
  library(sva)
})


# Set options
options(max.print = 1000, scipen = 999, stringsAsFactors = FALSE)

# Source helper functions
source(here("scr/modif_diann_adapter.R"))
source(here("scr/helper_functions.R"))

# Load shared pipeline configuration
config <- load_stage_parameters(
  stage = "exploratory",
  config_path = default_pipeline_config_path()
)

# Create results directories
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(rds_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r define_parameters}
#| cache: false
# Parameters are loaded from config.yaml in the setup chunk.
invisible(config)
```

# Data Loading and Annotation

```{r load_data}
#| cache: false
# Load data based on search engine
rds_path_data <- file.path(rds_dir, paste0(pre_fix, "_", search_engine, "_data.rds"))

if (file.exists(rds_path_data)) {

  message("Loading cached ", search_engine, " data...")
  df_from_search <- read_rds(rds_path_data)
  
} else {
  message("Loading ", search_engine, " data...")
  
  # Load data based on search engine type
  df_from_search <- switch(search_engine,
    "diann" = {
      # DIA-NN data loading
      file_ext <- tools::file_ext(diann_report_location)
      
      if (file_ext == "parquet") {
        message("Using parquet adapter for DIA-NN report.parquet file...")

        diann_adapter_parquet(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )
      } else if (file_ext == "tsv") {
        message("Using standard diann_adapter for DIA-NN report.tsv file...")
        result <- diann_adapter(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )

        # Clean column names for TSV files
        result %>% rename_with(~ str_remove(., "_.*$"), starts_with(instrument))

      } else {
        stop("Unsupported DIA-NN file format. Please use either .parquet or .tsv files.")
      }
    },
    
    "fragpipe_tmt" = {
      message("Using FragPipe TMT adapter...")
      fragpipe_adapter(
        parent_dir = fragpipe_parent_dir,
        ref_sample = ref_sample,
        grouping_var = "nterm_modif_peptide",
        min_purity = min_purity,
        tmt_delta = tmt_delta
      )
    },
    
    "fragpipe_lf" = {
      message("Using FragPipe Label-Free adapter...")
      fragpipe_lf_adapter(
        parent_dir = fragpipe_lf_parent_dir,
        annotation_file_path = fragpipe_lf_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "fragpipe_heavy_light" = {
      message("Using FragPipe Heavy-Light adapter...")
      fragpipe_dda_heavy_light_adapter(
        combined_modified_peptide_file = fragpipe_hl_file,
        annotation_file_path = fragpipe_hl_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "spectronaut" = {
      message("Using Spectronaut adapter...")
      spectronaut_adapter(
        path_to_file = spectronaut_report,
        proteotypic = proteotypic_only
      )
    },
    
    stop("Unsupported search engine: ", search_engine, 
         ". Supported options: diann, fragpipe_tmt, fragpipe_lf, fragpipe_heavy_light, spectronaut")
  )
  
  write_rds(df_from_search, rds_path_data)

}

# Load experimental design (adapt based on search engine
# For other search engines, load external annotation file
experimental_design <- load_pipeline_experimental_design(
  location_annotation = location_annotation,
  search_data_colnames = colnames(df_from_search),
  instrument = instrument
)

# Load protein to gene mapping with caching (aligned to data_prep_prelim outputs)
rds_path_prot2gene <- file.path(rds_dir, paste0(pre_fix, "_", search_engine, "_prot2gene.rds"))

if (file.exists(rds_path_prot2gene)) {

  message("Loading cached protein-to-gene mapping...")
  prot2gene <- read_rds(rds_path_prot2gene)

} else {

  prot2gene <- read_tsv(
    uniprot_annotation
  ) %>%
  janitor::clean_names() %>%
  dplyr::select(
    protein = entry,
    gene = gene_names_primary) %>%
  filter(
    protein %in% unique(df_from_search$protein)
  )

  write_rds(prot2gene, rds_path_prot2gene)
}
```

```{r annotate_peptides}
#| cache: false
# Annotate neo-termini
rds_path_annotated <- file.path(rds_dir, paste0(pre_fix, "_annotated_data.rds"))

if (file.exists(rds_path_annotated)) {

  message("Loading cached annotated data...")
  annotated_df_quant <- read_rds(rds_path_annotated)
  # Ensure flags exist even when loading from cache
  #annotated_df_quant <- add_interest_flags(annotated_df_quant)

} else {
  message("Annotating neo-termini...")

  annotated_df_quant <- annotate_neo_termini(
    peptides_df = df_from_search,
    fasta_location = fasta_location,
    sense = sense_protease,
    specificity = specificity_protease,
    organism = organism_annotation
  ) %>% 
    mutate(cleav_len = nchar(cleavage_sequence)) %>% 
    relocate(cleav_len, .after = processing_annotation_end) %>%
    filter(!is.na(peptide_start))

  annotated_df_quant <- annotated_df_quant %>%
    mutate(
      is_in_scope_for_interest = case_when(
        specificity %in% interesting_specificity &
          nterm_modif %in% interesting_nterm_modif ~ TRUE,
        TRUE ~ FALSE
      )
    )

  
  write_rds(annotated_df_quant, rds_path_annotated)
}
```

# Quick look at experimental annotation and samples

```{r}
experimental_design
```

```{r}
gtsummary::tbl_summary(
  experimental_design,
  by = condition)
```

# Correction of experimental metadata annonation

Some MS runs had to be excluded during the search because they seemed to be corrupted. The search crashed with those. This was tested after re-transferring the files and the search was tested in both Windows and Linux environments. Yielded the same results. We will exclude those from the experimental design here, to it matches the final list of samples in the search results.

If a patient has less than 2 replicates, we will exclude that patient entirely from the analysis.

```{r}
experimental_design <- filter_pipeline_experimental_design(
  experimental_design = experimental_design,
  observed_samples = colnames(annotated_df_quant),
  replicate_column = "replicate"
)
```

Keep only samples in the annotated_df_quant that are in the updated experimental_design

```{r}
samples_in_design <- experimental_design$sample

samples_searched <- colnames(annotated_df_quant) %>%
  .[str_starts(., instrument)]

# which samples should be excluded?
samples_to_exclude <- setdiff(samples_searched, samples_in_design)

annotated_df_quant <- annotated_df_quant %>%
  dplyr::select(-all_of(samples_to_exclude))
```

# Identification Statistics

```{r identification_stats}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Calculate identification statistics
create_identification_plots <- function(data, experimental_design) {

  # Number of identifications per sample
  ids_per_sample <- data %>%
    dplyr::select(starts_with(instrument)) %>%
    pivot_longer(cols = starts_with(instrument), names_to = "sample", values_to = "value") %>%
    group_by(sample) %>%
    summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
    left_join(experimental_design, by = "sample")
  
  # Plot identifications by condition
  p1 <- ids_per_sample %>%
    ggplot(aes(x = sample, y = n_identifications, fill = condition)) +
    geom_col(position = "dodge") +
    labs(title = "Number of Identifications by Condition",
         x = "Condition", y = "Number of Identifications") +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90, hjust = 1),
      )
  
  return(list(plot = p1, data = ids_per_sample))
}

id_results <- create_identification_plots(annotated_df_quant, experimental_design)
print(id_results$plot)
```

# Median Identifications by Processing Type

Performed the search only N-term semispecific to make it faster and because of our focus on N-terminal acetylation.

```{r}
annotated_df_quant$nterm_modif %>% table()
```

```{r median_ids_processing_type}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Calculate identifications by specificity and refined processing type (excluding specific peptides)
ids_by_specificity_refined <- annotated_df_quant %>%
  filter(specificity %in% c("semi_Cterm", "semi_Nterm")) %>%
  dplyr::select(specificity, processing_type, starts_with(instrument)) %>%
  pivot_longer(cols = starts_with(instrument),
               names_to = "sample",
               values_to = "value") %>%
  group_by(specificity, processing_type, sample) %>%
  summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
  left_join(
    experimental_design, 
    by = "sample")

# Calculate median identifications by refined processing type
median_ids_refined_processing <- ids_by_specificity_refined %>%
  group_by(specificity, processing_type, condition) %>%
  summarise(
    median_ids = median(n_identifications),
    mean_ids = mean(n_identifications),
    sd_ids = sd(n_identifications),
    n_samples = n(),
    .groups = 'drop'
  )

# Plot median identifications with facet_grid (horizontal bars)
plot_median_refined <- median_ids_refined_processing %>%
  ggplot(aes(x = median_ids, 
             y = condition,
             fill = condition)) +
  geom_col() +
  geom_errorbar(aes(xmin = median_ids - sd_ids, 
                    xmax = median_ids + sd_ids),
                width = 0.3, alpha = 0.7) +
  geom_text(aes(label = round(median_ids, 0)), 
            hjust = -0.01, vjust = -0.2, size = 2.5) +
  facet_grid(
    processing_type ~ specificity, 
    scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Number of Identifications by Refined Processing Type\nBefore Sparcity Reduction and Imputation",
    subtitle = "Semi-specific peptides only, grouped by specificity and processing type",
    x = "Median number of identifications",
    y = "Condition",
    fill = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 9)
  )

print(plot_median_refined)
```

# Median Identifications by Processing Type and N-term modification

Performed the search only N-term semispecific to make it faster and because of our focus on N-terminal acetylation.


```{r median_ids_processing_type_nterm_modif}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Calculate identifications by specificity and refined processing type (excluding specific peptides)
ids_by_specificity_refined <- annotated_df_quant %>%
  filter(specificity %in% c("semi_Cterm", "semi_Nterm")) %>%
  dplyr::select(specificity, processing_type, nterm_modif, starts_with(instrument)) %>%
  pivot_longer(cols = starts_with(instrument),
               names_to = "sample",
               values_to = "value") %>%
  group_by(specificity, nterm_modif, processing_type, sample) %>%
  summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
  left_join(
    experimental_design, 
    by = "sample")

# Calculate median identifications by refined processing type
median_ids_refined_processing <- ids_by_specificity_refined %>%
  group_by(specificity, nterm_modif, processing_type, condition) %>%
  summarise(
    median_ids = median(n_identifications),
    mean_ids = mean(n_identifications),
    sd_ids = sd(n_identifications),
    n_samples = n(),
    .groups = 'drop'
  )

# Plot median identifications with facet_grid (horizontal bars)
plot_median_refined <- median_ids_refined_processing %>%
  ggplot(aes(x = median_ids, 
             y = condition,
             fill = condition)) +
  geom_col() +
  geom_errorbar(aes(xmin = median_ids - sd_ids, 
                    xmax = median_ids + sd_ids),
                width = 0.3, alpha = 0.7) +
  geom_text(aes(label = round(median_ids, 0)), 
            hjust = -0.01, vjust = -0.2, size = 2.5) +
  facet_grid(
    processing_type ~ nterm_modif + specificity, 
    scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Number of Identifications by Refined Processing Type\nBefore Sparcity Reduction and Imputation",
    subtitle = "Semi-specific peptides only, grouped by specificity and processing type",
    x = "Median number of identifications",
    y = "Condition",
    fill = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 9)
  )

print(plot_median_refined)

# Display summary table
knitr::kable(median_ids_refined_processing, 
             caption = "Summary Statistics: Identifications by Refined Processing Type and Condition",
             digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Missing Values Analysis

```{r missing_values}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Analyze missing value patterns
missing_analysis <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  vis_miss(warn_large_data = FALSE) +
  labs(title = "Missing Value Patterns")

print(missing_analysis)

# Calculate missing value statistics
missing_stats <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "missing_count") %>%
  left_join(experimental_design, by = "sample")

missing_plot <- missing_stats %>%
  ggplot(aes(x = sample, y = missing_count, fill = condition)) +
  geom_col() +
  labs(title = "Missing Values by Sample", x = "Sample", y = "Missing Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(missing_plot)
```

# Imputation and Data Preparation

```{r imputation}
#| cache: false
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height 

# Prepare data for downstream analysis
rds_path_prepared <- file.path(rds_dir, paste0(pre_fix, "_prepared_n_imputed_data.rds"))
imputation_list_path <- file.path(rds_dir, paste0(pre_fix, "_imputation_list.rds"))

if (file.exists(rds_path_prepared)) {

  message("Loading cached prepared data...")
  se_pept_imp_raw <- read_rds(rds_path_prepared)

  if (file.exists(imputation_list_path)) {
    se_pept_imp_raw_imputation_list <- read_rds(imputation_list_path)
  }

} else {
  message("Preparing data...")
  
  # Create quantification matrix
  quant_matrix <- annotated_df_quant %>%
    dplyr::select(nterm_modif_peptide, starts_with(instrument)) %>%
    column_to_rownames("nterm_modif_peptide") %>%
    # make sure that the order of the columns in quant_matrix matches the order in sample experimental_design
    dplyr::select(all_of(experimental_design$sample)) %>%
    as.matrix()
  
  # For TMT data, work only with features present in all samples
  if (search_engine == "fragpipe_tmt") {

    message("TMT data detected: filtering for complete cases only...")
    complete_features <- complete.cases(quant_matrix)
    quant_matrix <- quant_matrix[complete_features, ]
    
    # Filter annotation data to match
    annotation_data <- annotated_df_quant %>%
      filter(nterm_modif_peptide %in% rownames(quant_matrix)) %>%
      # force the order of the rows to match the quant_matrix
      dplyr::arrange(match(nterm_modif_peptide, rownames(quant_matrix))) %>%
      dplyr::select(-starts_with(instrument))
    
    message("Retained ", nrow(quant_matrix), " complete features out of ", 
            nrow(annotated_df_quant), " total features")

  } else {

    # For other data types, perform imputation as before
    message("Non-TMT data: performing imputation...")
    annotation_data <- annotated_df_quant %>%
      dplyr::select(-starts_with(instrument))
  }
  
  # Create SummarizedExperiment object
  se_pept_imp_raw <- SummarizedExperiment(
    assays = list(counts = quant_matrix),
    colData = experimental_design,
    rowData = annotation_data
  )
  
  # Apply imputation only for non-TMT data
  if (search_engine != "fragpipe_tmt") {
    se_pept_imp_raw_imputation_list <- terminer_imputation(
      se = se_pept_imp_raw,
      min_fraction_condition = missing_accepted,
      tune_sigma = tune_sigma,
      tune_quantile = tune_quantile
    )
    
    se_pept_imp_raw <- se_pept_imp_raw_imputation_list$se

    # get summary table of missingness and imputation approach by feature observation
    imputation_summary_table <- se_pept_imp_raw_imputation_list$imputation_summary_table

    write_tsv(imputation_summary_table, 
          file.path(results_dir, paste0(pre_fix, "_imputation_summary_table.tsv")))

    # Visualization: reconstruct long imputed dataframe and plot summaries (conditional)
    if (exists("imputation_summary_table") && is.data.frame(imputation_summary_table) && nrow(imputation_summary_table) > 0) {
      message("Creating imputation diagnostic plots...")

      # Reconstruct long imputed dataframe from the imputed SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(cols = -nterm_modif_peptide, names_to = "sample", values_to = "Abundance") %>%
        dplyr::left_join(imputation_summary_table, by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      # Only plot if imputation method column present
      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        try({
          # Histogram by condition and imputation method
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          # Density plot by condition and imputation method
          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          # Barplot for counts by imputation method
          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)

        # Mark that imputation plots were generated in this session
        imputation_plots_generated <- TRUE

      } else {
        message("Column 'imputation_method' not found in imputation summary; skipping imputation-method visualizations.")
      }

    } else {
      message("No imputation summary available — skipping imputation visualizations.")
    }
  }
  
  write_rds(se_pept_imp_raw, rds_path_prepared)

  if (exists("se_pept_imp_raw_imputation_list") && is.list(se_pept_imp_raw_imputation_list)) {
    write_rds(se_pept_imp_raw_imputation_list, imputation_list_path)
  }
}
```

```{r imputation_plots_from_cache}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# If we loaded cached prepared data earlier, imputation plots may not have been created.
# Recreate them from the saved TSV if needed.
if (!exists("imputation_plots_generated") || !isTRUE(imputation_plots_generated)) {

  if (!isTRUE(run_imputation_diagnostics)) {
    message("Imputation diagnostics disabled (run_imputation_diagnostics = FALSE). Skipping cached imputation plots.")
  } else {

  imp_sum_path <- file.path(results_dir, paste0(pre_fix, "_imputation_summary_table.tsv"))
  
  if (file.exists(imp_sum_path)) {

    tsv_mb <- file.info(imp_sum_path)$size / (1024^2)
    if (is.finite(tsv_mb) && tsv_mb > imputation_tsv_max_mb) {
      message(
        "Imputation summary TSV is large (~", round(tsv_mb, 1), " MB). ",
        "Skipping cached imputation plots. Increase imputation_tsv_max_mb or set run_imputation_diagnostics = TRUE to force."
      )
    } else {
  
    imputation_summary_table_cached <- readr::read_tsv(imp_sum_path, show_col_types = FALSE)
    if (is.data.frame(imputation_summary_table_cached) && nrow(imputation_summary_table_cached) > 0) {
      # Reconstruct long imputed dataframe from the (possibly cached) SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(cols = -nterm_modif_peptide, names_to = "sample", values_to = "Abundance") %>%
        dplyr::left_join(imputation_summary_table_cached, by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        try({
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)
      }
    }
    }
  }
  }
}
```

# Principal Component Analysis

```{r pca_analysis_1}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

perform_pca_analysis <- function(se_object, title_suffix = "", subtitle = "Visualizing conditions", color_var = "condition") {
  pca_matrix <- assay(se_object) %>% t() %>% na.omit()
  
  pca_result <- prcomp(pca_matrix, scale. = TRUE)
  
  pca_df <- data.frame(
    sample = rownames(pca_result$x),
    PC1 = pca_result$x[,1],
    PC2 = pca_result$x[,2]
  ) %>%
    left_join(experimental_design, by = "sample")
  
  var_exp <- summary(pca_result)$importance[2,]
  
  pca_plot <- pca_df %>%
    ggplot(aes(x = PC1, y = PC2, color = .data[[color_var]])) +
    geom_point(size = 5) +
    #stat_ellipse(level = 0.68) +
    labs(
      title = paste("PCA Analysis", title_suffix),
      subtitle = subtitle,
      x = paste("PC1 -", round(var_exp[1] * 100, 1), "% variance"),
      y = paste("PC2 -", round(var_exp[2] * 100, 1), "% variance"),
      color = str_to_title(str_replace_all(color_var, "_", " "))
    ) +
    theme_minimal()
  
  return(list(plot = pca_plot, data = pca_df, variance = var_exp))
}

# Optional subsetting for PCA
se_for_pca <- if (isTRUE(pca_use_interest_only)) {
  keep <- rowData(se_pept_imp_raw)$is_in_scope_for_interest
  keep[is.na(keep)] <- FALSE
  se_pept_imp_raw[keep, , drop = FALSE]
} else {
  se_pept_imp_raw
}

pca_results_1 <- perform_pca_analysis(
  se_for_pca, 
  title_suffix = "- Before Batch Correction",
  subtitle = if (isTRUE(pca_use_interest_only)) "PCA on interesting features" else "Visualizing conditions",
  color_var = "condition"
)

print(pca_results_1$plot)
```

```{r batch_effect_pca}
pca_results_batch <- perform_pca_analysis(
  se_for_pca, 
  title_suffix = "- ",
  subtitle = if (isTRUE(pca_use_interest_only)) "PCA on interesting features" else "Visualizing batches",
  color_var = "measurement_batch"
)

print(pca_results_batch$plot)
```

# Batch effect correction with ComBat 

```{r}
#| cache: false

# Perform batch effect correction using ComBat only if requested by the user
if (isTRUE(do_batch_correction)) {

  # Validate that the expected batch column exists
  if (!"plex" %in% colnames(experimental_design)) {
    stop("do_batch_correction is TRUE but 'plex' column not found in experimental_design. Either set do_batch_correction <- FALSE or provide an appropriate batch column named 'plex'.")
  }

  message("Running ComBat batch correction using experimental_design$plex as batch groups")
  combat_corrected <- ComBat(
    dat = assay(se_pept_imp_raw),
    batch = experimental_design$plex,
    mod = model.matrix(~ experimental_design$condition)
  )

  # Create a new SummarizedExperiment with corrected data
  se_pept_imp_corrected <- SummarizedExperiment(
    assays = list(counts = combat_corrected),
    colData = experimental_design,
    rowData = rowData(se_pept_imp_raw)
  )

} else {
  message("do_batch_correction is FALSE — skipping ComBat. If you want to run ComBat, set do_batch_correction <- TRUE. Note: batch groups default to experimental_design$plex.")
  # Keep the raw imputed SE unchanged when skipping batch correction
  se_pept_imp_corrected <- se_pept_imp_raw
}
```

## PCA Analysis After Batch Correction

```{r pca_after_correction}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

if (isTRUE(do_batch_correction)) {

  # Optional subsetting for PCA on corrected data
  se_corr_for_pca <- if (isTRUE(pca_use_interest_only)) {
    keep <- rowData(se_pept_imp_corrected)$is_in_scope_for_interest
    keep[is.na(keep)] <- FALSE
    se_pept_imp_corrected[keep, , drop = FALSE]
  } else {
    se_pept_imp_corrected
  }

  pca_corrected_condition <- perform_pca_analysis(
    se_corr_for_pca, 
    title_suffix = "- After Batch Correction, non-protein-normalized",
    subtitle = if (isTRUE(pca_use_interest_only)) "PCA on interesting features" else "Visualizing conditions",
    color_var = "condition"
  )

  print(pca_corrected_condition$plot)

} else {
  message("Skipping PCA 'After Batch Correction' — batch correction disabled (do_batch_correction = FALSE).")
}
```

```{r batch_effect_correction}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

if (isTRUE(do_batch_correction)) {
  # Optional subsetting for PCA on corrected data (batch)
  se_corr_for_pca_batch <- if (isTRUE(pca_use_interest_only)) {
    keep <- rowData(se_pept_imp_corrected)$is_in_scope_for_interest
    keep[is.na(keep)] <- FALSE
    se_pept_imp_corrected[keep, , drop = FALSE]
  } else {
    se_pept_imp_corrected
  }

  pca_corrected_batch <- perform_pca_analysis(
    se_corr_for_pca_batch, 
    title_suffix = "- After Batch Correction\nnon-protein-normalized",
    subtitle = if (isTRUE(pca_use_interest_only)) "PCA on interesting features" else "Visualizing batch effects",
    color_var = "plex"
  )

  print(pca_corrected_batch$plot)

} else {
  message("Skipping PCA 'After Batch Correction (batch view)' — batch correction disabled (do_batch_correction = FALSE).")
}
```

# Protein-level normalization and SE creation

```{r protein_level_normalization}
#| cache: false
# Build protein-normalized peptide abundances and paired SummarizedExperiment objects
# This enables auditing of peptides excluded by protein-level normalization.

# 1) Build a long peptide table from corrected data with required annotations
#    Use non-log2 (linear) abundances for protein-level normalization
pept_q_raw <- (2 ^ assay(se_pept_imp_corrected)) %>%
  as.data.frame() %>%
  rownames_to_column("nterm_modif_peptide") %>%
  left_join(
    as.data.frame(rowData(se_pept_imp_corrected)) %>%
      dplyr::select(
        nterm_modif_peptide,
        peptide,
        protein,
        nterm_modif,
        specificity
      ),
    by = "nterm_modif_peptide"
  ) %>%
  dplyr::relocate(
    nterm_modif_peptide,
    peptide,
    protein,
    nterm_modif,
    specificity
  )

pept_q_raw_annot <- pept_q_raw %>%
  dplyr::select(
    nterm_modif_peptide,
    peptide,
    protein,
    nterm_modif,
    specificity
  )

# remove samples from experiment


# 2) Run peptide-to-protein normalization
#    Using experimental_design as the sample annotation
protein_normalized_peptides <- peptide2protein_normalization(
  peptides = pept_q_raw,
  annot = experimental_design,
  peptide_annot = pept_q_raw_annot,
  summarize_by_specificity = TRUE
)


# 3) Prepare wide matrix of protein-normalized peptide abundances
prot_norm_wide <- protein_normalized_peptides$protein_normalized_pepts_scaled

# Determine which columns correspond to samples
prefixed_cols <- paste0("fraction_int_peptide2prot_", experimental_design$sample)

sample_cols <- intersect(colnames(prot_norm_wide), prefixed_cols)

if (length(sample_cols) == 0) {
  # Fallback: maybe already de-prefixed
  sample_cols <- intersect(colnames(prot_norm_wide), experimental_design$sample)
}

mat_df <- prot_norm_wide %>%
  dplyr::select(nterm_modif_peptide, all_of(sample_cols))

# Remove prefix if present so column names match experimental_design$sample
colnames(mat_df) <- colnames(mat_df) %>%
  stringr::str_remove("^fraction_int_peptide2prot_")

# Convert to matrix and order columns to match design
mat_prot_norm <- mat_df %>%
  column_to_rownames("nterm_modif_peptide") %>%
  as.matrix()

common_samples <- intersect(colnames(mat_prot_norm), experimental_design$sample)

mat_prot_norm <- mat_prot_norm[, common_samples, drop = FALSE]

# Align colData to the matrix columns and set rownames to sample IDs
design_common <- experimental_design %>%
  dplyr::filter(sample %in% common_samples) %>%
  dplyr::arrange(match(sample, common_samples))

rownames(design_common) <- design_common$sample

# Ensure columns strictly follow design order
mat_prot_norm <- mat_prot_norm[, design_common$sample, drop = FALSE]

# 4) Build annotation for protein-normalized rows by merging back rowData
prot_norm_join <- mat_df %>%
  left_join(
    as.data.frame(rowData(se_pept_imp_corrected)),
    by = "nterm_modif_peptide"
  )

rowdata_cols <- setdiff(colnames(prot_norm_join), colnames(mat_df))

annot_prot_norm <- prot_norm_join %>%
  dplyr::select(nterm_modif_peptide, dplyr::all_of(rowdata_cols)) %>%
  distinct()

# Ensure annotation order matches matrix rows
annot_prot_norm <- annot_prot_norm %>%
  dplyr::filter(nterm_modif_peptide %in% rownames(mat_prot_norm)) %>%
  arrange(match(nterm_modif_peptide, rownames(mat_prot_norm))) %>%
  mutate(protein_normalization = "protein_normalized")
  
# Set rownames to match assay rownames
rownames(annot_prot_norm) <- annot_prot_norm$nterm_modif_peptide

# 5) Identify peptides lost by protein-level normalization
all_features <- rownames(assay(se_pept_imp_corrected))
lost_features <- setdiff(all_features, rownames(mat_prot_norm))

n_lost <- length(lost_features)
message("Protein-level normalization: lost ", n_lost, " peptides out of ", length(all_features))

# 6) For lost peptides, bring their original corrected intensities and annotate
quant_lost_df <- assay(se_pept_imp_corrected)[lost_features, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("nterm_modif_peptide")

annot_lost <- as.data.frame(rowData(se_pept_imp_corrected)) %>%
  dplyr::filter(nterm_modif_peptide %in% lost_features) %>%
  mutate(protein_normalization = "not_protein_normalized")

# 7) Combine protein-normalized peptides with the lost ones (kept as original corrected intensities)
prot_norm_df <- mat_prot_norm %>%
  as.data.frame() %>%
  rownames_to_column("nterm_modif_peptide")

combined_df <- bind_rows(prot_norm_df, quant_lost_df)

mat_prot_norm_with_lost <- combined_df %>%
  column_to_rownames("nterm_modif_peptide") %>%
  as.matrix()

annot_with_lost <- bind_rows(annot_prot_norm, annot_lost) %>%
  arrange(match(nterm_modif_peptide, rownames(mat_prot_norm_with_lost)))

# Constrain columns to design and set row/col alignment
mat_prot_norm_with_lost <- mat_prot_norm_with_lost[, design_common$sample, drop = FALSE]

rownames(annot_with_lost) <- annot_with_lost$nterm_modif_peptide

# 8) Create SE objects
se_pept_protnorm_pure <- SummarizedExperiment(
  assays = list(counts = mat_prot_norm),
  colData = design_common,
  rowData = annot_prot_norm
)

se_pept_protnorm_with_lost <- SummarizedExperiment(
  assays = list(counts = mat_prot_norm_with_lost),
  colData = design_common,
  rowData = annot_with_lost
)

# Non-protein-normalized SE aligned to the same sample set/order
counts_nonprot <- assay(se_pept_imp_corrected)[, design_common$sample, drop = FALSE]

rowdata_nonprot <- as.data.frame(rowData(se_pept_imp_corrected)) %>%
  mutate(protein_normalization = "none") %>%
  dplyr::filter(nterm_modif_peptide %in% rownames(counts_nonprot)) %>%
  arrange(match(nterm_modif_peptide, rownames(counts_nonprot)))

rownames(rowdata_nonprot) <- rowdata_nonprot$nterm_modif_peptide

se_pept_non_protnorm <- SummarizedExperiment(
  assays = list(counts = counts_nonprot),
  colData = design_common,
  rowData = rowdata_nonprot
)

# Clean up large temporaries
rm(pept_q_raw, pept_q_raw_annot, prot_norm_wide, mat_df, prot_norm_join,
   prot_norm_df, combined_df, quant_lost_df, annot_lost)
   
invisible(gc())
```

```{r}
# get protein quant
mat_prot_quant <- protein_normalized_peptides$summarized_protein_abundance_scaled %>%  
  dplyr::select(
    protein,
    matches("_prot")
  ) %>%
  column_to_rownames("protein") %>%
  as.matrix()

colnames(mat_prot_quant) <- str_remove(
  colnames(mat_prot_quant), 
  "_prot"
)

# get protein annotation
prot_quant_annot <- tibble(
  protein = rownames(mat_prot_quant)
) %>%
  left_join(., prot2gene, 
            by = "protein") 

# prepare summarized experiment object for protein-level quant
data_prot_quant_se <- SummarizedExperiment(
  assays = list(counts = mat_prot_quant),
  colData = design_common,
  rowData = prot_quant_annot
)
```

# Principal Component Analysis of Protein-Normalized Data

```{r pca_protein_norm_condition}
#| fig-width: !expr plot_width * side_by_side_scale
#| fig-height: !expr plot_height
#| cache: false

# Optional subsetting for PCA on protein-normalized (+ lost)
se_protnorm_for_pca <- if (isTRUE(pca_use_interest_only)) {

  keep <- rowData(se_pept_protnorm_with_lost)$is_in_scope_for_interest
  keep[is.na(keep)] <- FALSE
  se_pept_protnorm_with_lost[keep, , drop = FALSE]

} else {

  se_pept_protnorm_with_lost

}

pca_protnorm_with_lost <- perform_pca_analysis(
  se_protnorm_for_pca,
  title_suffix = "- Protein-normalized (+ lost)",
  subtitle = if (isTRUE(pca_use_interest_only)) "Visualizing conditions - interesting features" else "Visualizing conditions - all features",
  color_var = "condition"
)

# Reuse previously computed PCA for non-protein-normalized data
se_nonprot_for_pca <- if (isTRUE(pca_use_interest_only)) {
  keep <- rowData(se_pept_non_protnorm)$is_in_scope_for_interest
  keep[is.na(keep)] <- FALSE
  se_pept_non_protnorm[keep, , drop = FALSE]
} else {
  se_pept_non_protnorm
}

pca_non_protnorm <- perform_pca_analysis(
  se_nonprot_for_pca, 
  title_suffix = "- Non-protein-normalized",
  subtitle = if (isTRUE(pca_use_interest_only)) "Visualizing conditions - interesting features" else "Visualizing conditions - all features",
  color_var = "condition"
)

ggpubr::ggarrange(
  pca_protnorm_with_lost$plot,
  pca_non_protnorm$plot,
  ncol = 2,
  labels = c("A", "B")
)
```

```{r pca_protein_norm_intersest}
#| fig-width: !expr plot_width * side_by_side_scale
#| fig-height: !expr plot_height
#| cache: false

# Side-by-side PCA: all features vs only interesting features (protein-normalized + lost)
se_all <- se_pept_protnorm_with_lost

# "Interesting" uses the precomputed is_in_scope_for_interest flag
rd <- as.data.frame(rowData(se_all))

keep_int <- rd$is_in_scope_for_interest

keep_int[is.na(keep_int)] <- FALSE

n_int <- sum(keep_int)

if (n_int < 3) {
  message("Not enough interesting features for PCA (n = ", n_int, "). Showing only 'all features'.")
}

# Build SE for interesting features if enough rows
se_int <- if (n_int >= 3) se_all[keep_int, , drop = FALSE] else NULL

# Condition-colored PCAs
p_all_cond <- perform_pca_analysis(
  se_all,
  title_suffix = "- Protein-normalized (+lost)\nAll features",
  subtitle = "Color: condition",
  color_var = "condition"
)$plot

p_int_cond <- if (!is.null(se_int)) {
  perform_pca_analysis(
    se_int,
    title_suffix = "- Protein-normalized (+lost)\nInteresting features",
    subtitle = "Color: condition",
    color_var = "condition"
  )$plot
} else NULL

# Arrange side-by-side: All vs Interesting
plots_cond <- list(p_all_cond, p_int_cond)

plots_cond <- plots_cond[!vapply(plots_cond, is.null, logical(1))]

if (length(plots_cond) > 0) {
  print(ggpubr::ggarrange(plotlist = plots_cond, ncol = length(plots_cond), labels = c("A", "B")[seq_along(plots_cond)]))
}
```

# Save Results for Downstream Analysis

```{r save_results}
#| cache: false

# Save key objects for downstream analysis
save_objects <- list(
  annotated_df_quant = annotated_df_quant,
  # Keep the batch-corrected raw peptide SE for reference
  se_pept_imp_raw = se_pept_imp_corrected,
  # New objects for protein normalization auditing
  se_pept_protnorm_pure = se_pept_protnorm_pure,
  se_pept_protnorm_with_lost = se_pept_protnorm_with_lost,
  se_pept_non_protnorm = se_pept_non_protnorm,
  # summarized protein abundance
  se_prot_norm = data_prot_quant_se,
  lost_features = lost_features,
  experimental_design = experimental_design,
  prot2gene = prot2gene,
  parameters = list(
    search_engine = search_engine,
    sense_protease = sense_protease,
    specificity_protease = specificity_protease,
    organism_annotation = organism_annotation,
    missing_accepted = missing_accepted,
    pre_fix = pre_fix,
    interesting_specificity = interesting_specificity,
    interesting_nterm_modif = interesting_nterm_modif
  )
)

write_rds(save_objects, file.path(rds_dir, paste0(pre_fix, "_exploratory_results.rds")))

message("Exploratory analysis complete. Results saved for downstream analysis.")
```

# Session Information

```{r session_info}
sessionInfo()
```
