---
title: "TermineR Data Preparation"
author: "Your Name"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: false
    embed-resources: true
    df-print: paged
    max-width: "1200px"
editor: source
execute:
  eval: true
  echo: false
  error: false
  warning: true
  message: true
  cache: true
---

# Setup and Parameters

```{r setup}
#| cache: false
# Load required libraries
suppressPackageStartupMessages({
  library(TermineR)
  library(tidyverse)
  library(here)
  library(arrow)
  library(SummarizedExperiment)
  library(ggpubr)
  library(pheatmap)
  library(RColorBrewer)
  library(mixOmics)
  library(naniar)
  library(kableExtra)
  library(janitor)
  library(sva)
})


# Set options
options(max.print = 1000, scipen = 999, stringsAsFactors = FALSE)

# Source helper functions
source(here("scr/modif_diann_adapter.R"))
source(here("scr/helper_functions.R"))

# Create results directories
dir.create(here("results"), showWarnings = FALSE, recursive = TRUE)
dir.create(here("rds"), showWarnings = FALSE, recursive = TRUE)
rds_dir <- here("rds")
```

```{r define_parameters}
#| cache: false
# === SEARCH ENGINE CONFIGURATION ===
# EXPERIMENT-SPECIFIC: set the search engine used to produce your inputs.
search_engine <- "diann"  # Options: "diann", "fragpipe_tmt", "fragpipe_lf", "fragpipe_heavy_light", "spectronaut"

# === DATA LOCATIONS (adjust based on search engine) ===
# EXPERIMENT-SPECIFIC: update input data paths for your study.
# For DIA-NN
diann_report_location <- here("diann_report_or_parquet_location/.../report.tsv")  # Can be .parquet or .tsv

# For FragPipe TMT
fragpipe_parent_dir <- NULL #here("data/2024_05_semi_tryptic_search_fp21")  # Directory containing mix folders

# For FragPipe Label-Free
fragpipe_lf_parent_dir <- NULL #here("data/fp_semi")
fragpipe_lf_annotation <- NULL #here("data/OS001_ccRCC_experimental_annotation_current.txt")

# For FragPipe Heavy-Light
fragpipe_hl_file <- NULL #here("initial_data/combined_modified_peptide_label_quant.tsv")
fragpipe_hl_annotation <- NULL #here("initial_data/fragpipe_hl_annotation.txt")

# For Spectronaut
spectronaut_report <- NULL #here("initial_data/spectronaut_report.tsv")

# Common files
# EXPERIMENT-SPECIFIC: update annotation and proteome resources.
location_annotation <- here("data/XX000_experimental_metadata_file.txt")
fasta_location <- here("data/human_uniprot_2024_09/uniprotkb_proteome_UP000005640_AND_mode_2024_09_16.fasta")
targetp_location <- here("data/human_uniprot_2024_09/h_sapiens_long_summary.targetp2")  # Set to NULL if not available
uniprot_annotation <- here("data/uniprotkb_proteome_UP000005640_AND_mode_2025_08_13.tsv") # Uniprot annotation file

# === EXPERIMENTAL PARAMETERS ===
# EXPERIMENT-SPECIFIC: protease, organism, and sample naming conventions.
sense_protease <- "C"  # "C" for C-terminal cleavage (trypsin), "N" for N-terminal
specificity_protease <- "K|R"  # Trypsin specificity
organism_annotation <- "human"  # Options: "human", "mouse", "arabidopsis", etc.
instrument <- "TF"  # Instrument prefix in sample names

# === SEARCH ENGINE SPECIFIC PARAMETERS ===
# EXPERIMENT-SPECIFIC: set parameters according to your search engine outputs.
# FragPipe TMT parameters
ref_sample <- "norm"  # Reference channel for TMT normalization
min_purity <- 0.5  # Minimum PSM purity
tmt_delta <- "229"  # TMT delta mass ("229" for TMT10/11, "304" for TMT16)

# DIA-NN parameters
proteotypic_only <- TRUE  # Use only proteotypic peptides
summarization_method <- "SUM"  # "SUM" or "MAX" for precursor summarization

# === ANALYSIS PARAMETERS ===
# EXPERIMENT-SPECIFIC: missingness tolerance and output prefix.
missing_accepted <- 2 / 4  # Max missing values per condition (2 out of 4 replicates)
tune_sigma <- 1
tune_quantile <- 0.0000001
pre_fix <- "xx000_projectname_analysis_type"  # Prefix for output files

# === PLOT PARAMETERS ===
# EXPERIMENT-SPECIFIC: adjust plot dimensions for your report.
plot_width <- 10
plot_height <- 5
side_by_side_scale <- 2.0  # wider figures when arranging multiple plots
plot_dpi <- 300

# EXPERIMENT-SPECIFIC: define 'interesting' feature scope for PCA/DA.
# Interest scope for PCA
interesting_specificity <- c("semi_Nterm", "semi_Cterm")  # e.g., c("semi_Nterm", "semi_Cterm")
interesting_nterm_modif <- c("n", "Acetyl")  # e.g., c("Dimethyl", "Acetyl")

# EXPERIMENT-SPECIFIC: toggle optional workflows.
# Toggle: if TRUE, PCA will use only 'interesting' features (per add_interest_flags)
pca_use_interest_only <- FALSE

# Toggle: run ComBat batch-correction? Set to TRUE to perform ComBat. Note: batch groups default to experimental_design$plex
do_batch_correction <- FALSE
```

# Data Loading and Annotation

```{r load_data}
#| cache: true
#| message: true
# Load data based on search engine
rds_path_data <- file.path(rds_dir, paste0("pre_search_", search_engine, "_data.rds"))

if (file.exists(rds_path_data)) {

  message("Loading cached ", search_engine, " data...")
  df_from_search <- read_rds(rds_path_data)
  
} else {
  message("Loading ", search_engine, " data...")
  
  # Load data based on search engine type
  df_from_search <- switch(search_engine,
    "diann" = {
      # DIA-NN data loading
      file_ext <- tools::file_ext(diann_report_location)
      
      if (file_ext == "parquet") {
        message("Using parquet adapter for DIA-NN report.parquet file...")

        diann_adapter_parquet(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )
      } else if (file_ext == "tsv") {
        message("Using standard diann_adapter for DIA-NN report.tsv file...")
        result <- diann_adapter(
          path_to_file = diann_report_location,
          proteotypic = proteotypic_only,
          summarization = summarization_method
        )

        # Clean column names for TSV files
        result %>% rename_with(~ str_remove(., "_.*$"), starts_with(instrument))

      } else {
        stop("Unsupported DIA-NN file format. Please use either .parquet or .tsv files.")
      }
    },
    
    "fragpipe_tmt" = {
      message("Using FragPipe TMT adapter...")
      fragpipe_adapter(
        parent_dir = fragpipe_parent_dir,
        ref_sample = ref_sample,
        grouping_var = "nterm_modif_peptide",
        min_purity = min_purity,
        tmt_delta = tmt_delta
      )
    },
    
    "fragpipe_lf" = {
      message("Using FragPipe Label-Free adapter...")
      fragpipe_lf_adapter(
        parent_dir = fragpipe_lf_parent_dir,
        annotation_file_path = fragpipe_lf_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "fragpipe_heavy_light" = {
      message("Using FragPipe Heavy-Light adapter...")
      fragpipe_dda_heavy_light_adapter(
        combined_modified_peptide_file = fragpipe_hl_file,
        annotation_file_path = fragpipe_hl_annotation,
        grouping_var = "nterm_modif_peptide"
      )
    },
    
    "spectronaut" = {
      message("Using Spectronaut adapter...")
      spectronaut_adapter(
        path_to_file = spectronaut_report,
        proteotypic = proteotypic_only
      )
    },
    
    stop("Unsupported search engine: ", search_engine, 
         ". Supported options: diann, fragpipe_tmt, fragpipe_lf, fragpipe_heavy_light, spectronaut")
  )
  
  write_rds(df_from_search, rds_path_data)

}

# Load experimental design (adapt based on search engine
# For other search engines, load external annotation file
experimental_design <- read_delim(location_annotation) %>%
    mutate(
      measurement_batch = as.factor(measurement_batch)
    )

# Load protein to gene mapping with caching
rds_path_prot2gene <- file.path(rds_dir, paste0("pre_search_", search_engine, "_prot2gene.rds"))

if (file.exists(rds_path_prot2gene)) {

  message("Loading cached protein-to-gene mapping...")
  prot2gene <- read_rds(rds_path_prot2gene)

} else {

  prot2gene <- read_tsv(
    uniprot_annotation
  ) %>%
  janitor::clean_names() %>%
  dplyr::select(
    protein = entry,
    gene = gene_names_primary) %>%
  filter(
    protein %in% unique(df_from_search$protein)
  )

  write_rds(prot2gene, rds_path_prot2gene)
}
```

```{r annotate_peptides}
#| cache: false
# Annotate neo-termini
rds_path_annotated <- file.path(rds_dir, "pre_search_annotated_data.rds")

if (file.exists(rds_path_annotated)) {

  message("Loading cached annotated data...")
  annotated_df_quant <- read_rds(rds_path_annotated)
  # Ensure flags exist even when loading from cache
  #annotated_df_quant <- add_interest_flags(annotated_df_quant)

} else {
  message("Annotating neo-termini...")

  annotated_df_quant <- annotate_neo_termini(
    peptides_df = df_from_search,
    fasta_location = fasta_location,
    sense = sense_protease,
    specificity = specificity_protease,
    organism = organism_annotation
  ) %>% 
    mutate(cleav_len = nchar(cleavage_sequence)) %>% 
    relocate(cleav_len, .after = processing_annotation_end) %>%
    filter(!is.na(peptide_start))
  
  # Add TargetP annotation if available
  if (is.null(targetp_location)) {

    message("TargetP annotation not defined")
  } else {
    message("TargetP annotation defined. Loading")

  targetp_annotation_df <- read_tsv(targetp_location,
                               skip = 1) %>%
  clean_names() %>%
  separate(
    number_id,
    into = c("trembl", "protein", "uniprot_accession"),
    sep = "\\|",
    remove = FALSE
  ) %>%
  dplyr::rename(
    protein_header = number_id
  ) %>%
  mutate(
    targetp_p1_position = str_remove(cs_position,
                                     "CS pos:")
  ) %>%
  mutate(
    targetp_p1_position = str_remove(targetp_p1_position,
                                     # everything after the first "-"
                                     "-.*")
  ) %>%
  mutate(
    targetp_p1_position = as.numeric(targetp_p1_position)
  ) %>%
  filter(
    !is.na(targetp_p1_position)
  ) %>% 
  dplyr::select(
    protein,
    targetp_category = prediction,
    targetp_p1_position
  ) 
    
  annotated_df_quant <- annotated_df_quant %>%
    left_join(., targetp_annotation_df) %>%
    relocate(
      cleav_len,
      .after = cleavage_sequence
    ) %>%
    mutate(
      targetp_matches_p1_prime = case_when(
        abs(as.numeric(p1_position) - targetp_p1_position) < 4 ~ TRUE,
          TRUE ~ FALSE
      )
    ) %>%
    mutate(
      processing_type_targetp = case_when(
        targetp_matches_p1_prime == TRUE & targetp_category == "SP" ~ "SIGNAL",
        targetp_matches_p1_prime == TRUE & targetp_category == "mTP" ~ "mTP",
        targetp_matches_p1_prime == TRUE & targetp_category == "cTP" ~ "cTP",
        targetp_matches_p1_prime == TRUE & targetp_category == "luTP" ~ "luTP",
        targetp_matches_p1_prime == FALSE ~ "not_canonical",
        is.na(targetp_matches_p1_prime) ~ "not_canonical",
        TRUE ~ "not_canonical"
      )) %>%
    mutate(
      processing_type_terminer = dplyr::case_when(
        processing_type_targetp != "not_canonical" ~ processing_type_targetp,
        TRUE ~ dplyr::coalesce(uniprot_processing_type, "not_canonical_no_procc_annot")
      )
    ) %>%
    mutate(
      source_processing_annotation = case_when(
        (uniprot_processing_type == processing_type_targetp) & (uniprot_processing_type == "not_canonical") ~ "Other",
        (uniprot_processing_type == processing_type_targetp) & (processing_type_targetp == "not_canonical") ~ "Other",
        (uniprot_processing_type == processing_type_targetp) & (uniprot_processing_type == "not_canonical_no_procc_annot") ~ "Other",
        (uniprot_processing_type != processing_type_targetp) &  (uniprot_processing_type == "not_canonical") ~ "TargetP",
        (uniprot_processing_type != processing_type_targetp) &  (processing_type_targetp == "not_canonical") ~ "Uniprot",
        (uniprot_processing_type != processing_type_targetp) &  (uniprot_processing_type == "not_canonical_no_procc_annot") ~ "Other",
        (uniprot_processing_type != processing_type_targetp) &  (uniprot_processing_type != "not_canonical") ~ "Conflict/both",
        (uniprot_processing_type != processing_type_targetp) &  (uniprot_processing_type != "not_canonical_no_procc_annot") ~ "Conflict/both",
        (uniprot_processing_type != processing_type_targetp) &  (processing_type_targetp != "not_canonical") ~ "Conflict/both",
        TRUE ~ "not_canonical"
      )
    ) %>%
    mutate(
      source_processing_annotation = case_when(
        str_detect(processing_type_terminer, "not_canonical") ~ "Other", 
        TRUE ~ source_processing_annotation
      )) %>%
    mutate(
      processing_type = case_when(
        str_detect(processing_type_terminer, "INIT_MET") ~ "P2",
        processing_type_terminer == "Intact_ORF" ~ "P1",
        processing_type_terminer == "mTP" ~ "MTS", 
        processing_type_terminer == "SIGNAL" ~ "SP",
        processing_type_terminer == "not_canonical_no_procc_annot" ~ "not_canonical",
        TRUE ~ processing_type_terminer
      )
    ) %>%
    mutate(
      difference = case_when(
        targetp_matches_p1_prime == TRUE ~ p1_position - targetp_p1_position,
        matches_p1_prime == TRUE & targetp_matches_p1_prime == FALSE ~ p1_position - processing_annotation_end
      )
    ) %>%
    mutate(
      processing_type = factor(
        processing_type,
        levels = c("P1", "P2", "SP", "MTS", "cTP", "TRANSIT", "PROPEP", "not_canonical")
      )
    ) %>%
    relocate(targetp_category,
             targetp_p1_position,
             targetp_matches_p1_prime,
             processing_type_targetp,
             processing_type_terminer,
             processing_type,
             source_processing_annotation,
             .after = processing_annotation_end) %>%
    mutate(
      is_in_scope_for_interest = case_when(
        specificity %in% interesting_specificity &
          nterm_modif %in% interesting_nterm_modif ~ TRUE,
        TRUE ~ FALSE
      )
    )
  
  }  # Close 'else' for targetp_location

  
  write_rds(annotated_df_quant, rds_path_annotated)
}
```

# Quick look at experimental annotation and samples

```{r}
experimental_design
```

```{r}
gtsummary::tbl_summary(
  experimental_design,
  by = condition)
```

# Correction of experimental metadata annonation

Some MS runs had to be excluded during the search because they seemed to be corrupted. The search crashed with those. This was tested after re-transferring the files and the search was tested in both Windows and Linux environments. Yielded the same results. We will exclude those from the experimental design here, to it matches the final list of samples in the search results.

If a patient has less than 2 replicates, we will exclude that patient entirely from the analysis.

```{r}
experimental_design <- experimental_design %>%
  filter(sample %in% colnames(annotated_df_quant)) %>%
  group_by(bio_replicate) %>%
  filter(n() >= 2) %>%
  ungroup()
```

Keep only samples in the annotated_df_quant that are in the updated experimental_design

```{r}
samples_in_design <- experimental_design$sample

samples_searched <- colnames(annotated_df_quant) %>%
  .[str_starts(., instrument)]

# which samples should be excluded?
samples_to_exclude <- setdiff(samples_searched, samples_in_design)

annotated_df_quant <- annotated_df_quant %>%
  dplyr::select(-all_of(samples_to_exclude))
```

# Identification Statistics

```{r identification_stats}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Calculate identification statistics
create_identification_plots <- function(data, experimental_design) {

  # Number of identifications per sample
  ids_per_sample <- data %>%
    dplyr::select(starts_with(instrument)) %>%
    pivot_longer(cols = starts_with(instrument), names_to = "sample", values_to = "value") %>%
    group_by(sample) %>%
    summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
    left_join(experimental_design, by = "sample")
  
  # Plot identifications by condition
  p1 <- ids_per_sample %>%
    ggplot(aes(x = sample, y = n_identifications, fill = condition)) +
    geom_col(position = "dodge") +
    labs(title = "Number of Identifications by Condition",
         x = "Condition", y = "Number of Identifications") +
    theme_minimal() +
    theme(
      legend.position = "bottom",
      axis.text.x = element_text(angle = 90, hjust = 1),
      )
  
  return(list(plot = p1, data = ids_per_sample))
}

id_results <- create_identification_plots(annotated_df_quant, experimental_design)
print(id_results$plot)
```

# Median Identifications by Processing Type

Performed the search only N-term semispecific to make it faster and because of our focus on N-terminal acetylation.

```{r}
annotated_df_quant$nterm_modif %>% table()
```

```{r median_ids_processing_type}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Calculate identifications by specificity and refined processing type (excluding specific peptides)
ids_by_specificity_refined <- annotated_df_quant %>%
  filter(specificity %in% c("semi_Cterm", "semi_Nterm")) %>%
  dplyr::select(specificity, processing_type, nterm_modif, starts_with(instrument)) %>%
  pivot_longer(cols = starts_with(instrument),
               names_to = "sample",
               values_to = "value") %>%
  group_by(specificity, nterm_modif, processing_type, sample) %>%
  summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
  left_join(
    experimental_design, 
    by = "sample")

# Calculate median identifications by refined processing type
median_ids_refined_processing <- ids_by_specificity_refined %>%
  group_by(specificity, nterm_modif, processing_type, condition) %>%
  summarise(
    median_ids = median(n_identifications),
    mean_ids = mean(n_identifications),
    sd_ids = sd(n_identifications),
    n_samples = n(),
    .groups = 'drop'
  )

# Plot median identifications with facet_grid (horizontal bars)
plot_median_refined <- median_ids_refined_processing %>%
  ggplot(aes(x = median_ids, 
             y = condition,
             fill = condition)) +
  geom_col() +
  geom_errorbar(aes(xmin = median_ids - sd_ids, 
                    xmax = median_ids + sd_ids),
                width = 0.3, alpha = 0.7) +
  geom_text(aes(label = round(median_ids, 0)), 
            hjust = -0.01, vjust = -0.2, size = 2.5) +
  facet_grid(
    processing_type ~ nterm_modif + specificity, 
    scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Median Number of Identifications by Refined Processing Type\nBefore Sparcity Reduction and Imputation",
    subtitle = "Semi-specific peptides only, grouped by specificity and processing type",
    x = "Median number of identifications",
    y = "Condition",
    fill = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 9)
  )

print(plot_median_refined)

# Display summary table
knitr::kable(median_ids_refined_processing, 
             caption = "Summary Statistics: Identifications by Refined Processing Type and Condition",
             digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Missing Values Analysis

```{r missing_values}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# Analyze missing value patterns
missing_analysis <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  vis_miss(warn_large_data = FALSE) +
  labs(title = "Missing Value Patterns")

# Calculate missing value statistics
missing_stats <- annotated_df_quant %>%
  dplyr::select(starts_with(instrument)) %>%
  summarise_all(~sum(is.na(.))) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "missing_count") %>%
  left_join(experimental_design, by = "sample")

missing_plot <- missing_stats %>%
  ggplot(aes(x = sample, y = missing_count, fill = condition)) +
  geom_col() +
  labs(
    title = "Missing Values by Sample", 
    x = "Sample", 
    y = "Missing Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(missing_plot)
```

# Imputation and Data Preparation

```{r imputation}
#| cache: false
#| message: true
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height 

# Prepare data for downstream analysis
rds_path_prepared <- file.path(rds_dir, "pre_search_prepared_n_imputed_data.rds")

if (file.exists(rds_path_prepared)) {

  message("Loading cached prepared data...")
  se_pept_imp_raw <- read_rds(rds_path_prepared)

} else {
  message("Preparing data...")
  
  # Create quantification matrix
  quant_matrix <- annotated_df_quant %>%
    dplyr::select(nterm_modif_peptide, starts_with(instrument)) %>%
    column_to_rownames("nterm_modif_peptide") %>%
    # make sure that the order of the columns in quant_matrix matches the order in sample experimental_design
    dplyr::select(all_of(experimental_design$sample)) %>%
    as.matrix()
  
  # For TMT data, work only with features present in all samples
  if (search_engine == "fragpipe_tmt") {

    message("TMT data detected: filtering for complete cases only...")
    complete_features <- complete.cases(quant_matrix)
    quant_matrix <- quant_matrix[complete_features, ]
    
    # Filter annotation data to match
    annotation_data <- annotated_df_quant %>%
      filter(nterm_modif_peptide %in% rownames(quant_matrix)) %>%
      # force the order of the rows to match the quant_matrix
      dplyr::arrange(match(nterm_modif_peptide, rownames(quant_matrix))) %>%
      dplyr::select(-starts_with(instrument))
    
    message("Retained ", nrow(quant_matrix), " complete features out of ", 
            nrow(annotated_df_quant), " total features")

    # Create SummarizedExperiment object
    se_pept_imp_raw <- SummarizedExperiment(
      assays = list(counts = quant_matrix),
      colData = experimental_design,
      rowData = annotation_data
    )

  } else {

    # Prepare annotation data for non-TMT workflows
    annotation_data <- annotated_df_quant %>%
      dplyr::select(-starts_with(instrument))

    # Create SummarizedExperiment object
    se_pept_raw <- SummarizedExperiment(
      assays = list(counts = quant_matrix),
      colData = experimental_design,
      rowData = annotation_data
    )

    # For other data types, perform imputation as before
    message("Non-TMT data: performing imputation...")

    se_pept_imp_raw_imputation_list <- terminer_imputation(
      se = se_pept_raw,
      min_fraction_condition = missing_accepted,
      tune_sigma = 1,
      tune_quantile = 1e-7
    )

    se_pept_imp_raw <- se_pept_imp_raw_imputation_list$se

    # get summary table of missingness and imputation approach by feature observation
    imputation_summary_table <- se_pept_imp_raw_imputation_list$imputation_summary_table

    write_tsv(imputation_summary_table, 
               file.path(here("results"), paste0(pre_fix, "imputation_summary_table.tsv")))

    # Visualization: reconstruct long imputed dataframe and plot summaries (conditional)
    if (exists("imputation_summary_table") && is.data.frame(imputation_summary_table) && nrow(imputation_summary_table) > 0) {
      message("Creating imputation diagnostic plots...")

      # Reconstruct long imputed dataframe from the imputed SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(
          cols = -nterm_modif_peptide,
           names_to = "sample", 
           values_to = "Abundance") %>%
        dplyr::left_join(
          imputation_summary_table, 
          by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      # Only plot if imputation method column present
      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        message("Creating imputation diagnostic plots...")
        try({
          # Histogram by condition and imputation method
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          # Density plot by condition and imputation method
          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          # Barplot for counts by imputation method
          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)

        # Mark that imputation plots were generated in this session
        imputation_plots_generated <- TRUE

      } else {
        message("Column 'imputation_method' not found in imputation summary; skipping imputation-method visualizations.")
      }

    } else {
      message("No imputation summary available — skipping imputation visualizations.")
    }

  }  # end non-TMT branch
  
  write_rds(se_pept_imp_raw, rds_path_prepared)
  write_rds(se_pept_imp_raw_imputation_list, file.path(here("rds"), paste0(pre_fix, "_imputation_list.rds")))
}
```

```{r}
assay(se_pept_imp_raw)[1:10, 1:10]
```

```{r imputation_plots_from_cache}
#| fig-width: !expr plot_width
#| fig-height: !expr plot_height
#| cache: false

# If we loaded cached prepared data earlier, imputation plots may not have been created.
# Recreate them from the saved TSV if needed.
if (!exists("imputation_plots_generated") || !isTRUE(imputation_plots_generated)) {

  imp_sum_path <- file.path(here("results"), paste0(pre_fix, "imputation_summary_table.tsv"))
  
  if (file.exists(imp_sum_path)) {
  
    imputation_summary_table_cached <- readr::read_tsv(imp_sum_path, show_col_types = FALSE)
    if (is.data.frame(imputation_summary_table_cached) && nrow(imputation_summary_table_cached) > 0) {
      # Reconstruct long imputed dataframe from the (possibly cached) SummarizedExperiment
      imputed_mat_df <- as.data.frame(assay(se_pept_imp_raw)) %>%
        tibble::rownames_to_column(var = "nterm_modif_peptide")

      df_example_partial_missing_imp <- imputed_mat_df %>%
        tidyr::pivot_longer(cols = -nterm_modif_peptide, names_to = "sample", values_to = "Abundance") %>%
        dplyr::left_join(imputation_summary_table_cached, by = c("nterm_modif_peptide", "sample")) %>%
        dplyr::left_join(experimental_design, by = "sample")

      if ("imputation_method" %in% colnames(df_example_partial_missing_imp)) {
        try({
          p_hist_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_histogram(binwidth = 0.2, alpha = 0.5, position = "identity") +
            labs(
              title = "Distribution of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_hist_imp)

          p_dens_imp <- df_example_partial_missing_imp %>%
            ggplot(aes(x = Abundance, fill = imputation_method)) +
            geom_density(aes(y = after_stat(count)), alpha = 0.3) +
            labs(
              title = "Density of abundances by imputation approach",
              x = "Abundance",
              fill = "Imputation"
            ) +
            facet_grid(~condition) +
            theme(
              axis.text.x = element_text(size = 10),
              axis.text.y = element_text(size = 10),
              panel.background = element_blank(),
              panel.grid.major = element_line(color = "grey"),
              panel.border = element_rect(fill = NA, linewidth = 1.5),
              axis.title = element_text(size = 12, face = "bold")
            )

          print(p_dens_imp)

          p_bar_imp <- df_example_partial_missing_imp %>%
            dplyr::select(nterm_modif_peptide, sample, imputation_method) %>%
            distinct() %>%
            ggplot(aes(x = imputation_method, fill = imputation_method)) +
            geom_bar(fill = "#47A8BD") +
            geom_text(aes(label = after_stat(count)), stat = "count", vjust = 0.5, size = 3) +
            labs(title = "Nr of quantitative observations by type of imputation",
                 subtitle = "After exclusion of sparse features",
                 x = "Imputation type",
                 y = "Nr of quantitative observations") +
            theme_minimal() +
            theme(
              axis.text.x = element_text(size = 10, angle = 0, hjust = 0.5),
              axis.text.y = element_text(size = 10),
              axis.title = element_text(size = 12, face = "bold"),
              legend.position = "bottom"
            )

          print(p_bar_imp)

        }, silent = TRUE)
      }
    }
  }
}
```
