---
title: "TermineR Results Visualization"
author: "Your Name"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-depth: 3
    number-sections: false
    embed-resources: true
    df-print: paged
    max-width: "1200px"
editor: source
execute:
  eval: true
  echo: false
  error: false
  warning: false
  message: false
  cache: true
---
# Setup and Load Results

```{r setup}
#| cache: false
# Load required libraries
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(ggpubr)
  library(pheatmap)
  library(RColorBrewer)
  library(ComplexHeatmap)
  library(mixOmics)
  # Added for GO visualization
  library(clusterProfiler)
  library(org.Hs.eg.db)
  # Added for cleavage utilities
  library(TermineR)
  # Added for sequence logos
  library(dagLogo)
  library(SummarizedExperiment)
})

# Avoid lazy-load cache buffer issues in large objects
knitr::opts_chunk$set(cache.lazy = FALSE)

# Set parameters
rds_dir <- here("rds")
results_dir <- here("results")
plot_width <- 12
plot_height <- 8
# EXPERIMENT-SPECIFIC: update thresholds for your study.
fc_threshold <- 1.2
pval_threshold <- 0.05

# EXPERIMENT-SPECIFIC: update prefix used for inputs/outputs from other scripts.
pre_fix <- "xx000_projectname_analysis_type"

# EXPERIMENT-SPECIFIC: set protein annotation path (Uniprot TSV table of the proteome used for the search).
protein_annotation_path <- "Z:/xxxx/xxxx/xxxx/.../uniprotkb_proteome_UP000005640_AND_mode_2025_08_13.tsv"

# EXPERIMENT-SPECIFIC: select normalization dataset(s) of interest.
# Options available in this notebook environment: "protnorm_pure", "protnorm_with_lost", "non_protnorm", "protein_abund_norm" refers to protein-level analysis
dataset_of_interest <- c("protnorm_with_lost", "non_protnorm", "protein_abund_norm")

# EXPERIMENT-SPECIFIC: define condition levels in your experimental design.

condition_levels <- c(
  "NAT", 
  "PC",
  "Tumor"
)

# EXPERIMENT-SPECIFIC: define contrast levels matching your contrasts in inferential analysis.

contrast_levels <- c(
  "Tumor_v_NAT", 
  "PC_v_NAT",
  "Tumor_v_PC"
)

# EXPERIMENT-SPECIFIC: set sequence logo options (dagLogo).
run_sequence_logos <- FALSE
# Optional filters (NULL means all). Use names from `contrast_levels` and
# detected values in volcano_data_after_w_go$nterm_modif.
sequence_logo_contrasts <- NULL
sequence_logo_nterm_modifs <- NULL
```

```{r load_data}
#| cache: false
# Load pre-computed results
exploratory_results <- read_rds(file.path(rds_dir, paste0(pre_fix, "_exploratory_results.rds")))
limma_results <- read_rds(file.path(rds_dir, paste0(pre_fix, "_limma_results.rds")))

# Extract key objects
experimental_design <- exploratory_results$experimental_design

all_results <- limma_results$all_results %>%
  filter(dataset %in% dataset_of_interest)

params <- exploratory_results$parameters

pre_fix <- params$pre_fix

# New: SE with protein-normalized peptides plus lost ones, and the non-protein-normalized SE
se_with_lost <- exploratory_results$se_pept_protnorm_with_lost
se_non_protnorm <- exploratory_results$se_pept_non_protnorm
lost_features <- exploratory_results$lost_features

# protein to gene mapping
prot2gene <- exploratory_results$prot2gene

# Load GO enrichment results produced by the inferential analysis (if available)
go_results_path <- file.path(rds_dir, paste0(pre_fix, "_go_enrichment_results.rds"))

if (file.exists(go_results_path)) {

  message("Loading pre-computed GO enrichment results...")
  go_results_list <- read_rds(go_results_path)
  
  } else {
  
  message("No GO enrichment results found; skipping GO visualization.")
  go_results_list <- list()

}

# load go grouping table

go_groups_path <- file.path(here(rds_dir), paste0(pre_fix, "_go_cc_mapping.rds"))

if (file.exists(go_groups_path)) {

  message("Loading pre-computed GO grouping table...")
  go_groups_df <- read_rds(go_groups_path)
  
  } else {
  
  message("No GO grouping table found; skipping GO grouping visualization.")
  go_groups_df <- NULL

}
```

```{r}
# protein annotation
protein_annotation <- read_tsv(
  protein_annotation_path
) %>%
  janitor::clean_names() %>%
  dplyr::select(
    protein = entry,
    gene = gene_names_primary,
    protein_description = protein_names)
```

# Basic Identification Statistics

```{r}
# calculate number of identifications per sample
tabular_nr_of_ids_per_sample <- exploratory_results$annotated_df_quant %>%
  dplyr::select(all_of(exploratory_results$experimental_design$sample)) %>%
  # pivot_longer to get the data in a long format
  tidyr::pivot_longer(cols = all_of(exploratory_results$experimental_design$sample),
                      names_to = "sample",
                      values_to = "value") %>%
  group_by(sample) %>%
  summarise(
    n_ids_per_sample = sum(!is.na(value))
  ) 


# calculate number of identifications per sample and modification
tabular_nr_of_ids_per_sample_modif <- exploratory_results$annotated_df_quant %>%
  dplyr::select(nterm_modif, all_of(exploratory_results$experimental_design$sample)) %>%
  # pivot_longer to get the data in a long format
  tidyr::pivot_longer(cols = all_of(exploratory_results$experimental_design$sample),
                      names_to = "sample",
                      values_to = "value") %>%
  group_by(nterm_modif, sample) %>%
  summarise(
    n_ids_per_sample_n_modif = sum(!is.na(value))
  )

tabular_nr_of_ids_per_aa_before <- exploratory_results$annotated_df_quant %>%
  dplyr::select(
    nterm_modif_peptide, nterm_modif, p1_residue, 
    all_of(exploratory_results$experimental_design$sample)) %>%
  # pivot_longer to get the data in a long format
  tidyr::pivot_longer(cols = all_of(exploratory_results$experimental_design$sample),
                      names_to = "sample",
                      values_to = "value") %>%
  group_by(nterm_modif, p1_residue, sample) %>%
  summarise(
    n_identifications = sum(!is.na(value))
  ) %>%
  ungroup() %>%
  left_join(., tabular_nr_of_ids_per_sample) %>%
  left_join(., tabular_nr_of_ids_per_sample_modif) %>%
  mutate(
    percentage_of_ids_per_sample = n_identifications / n_ids_per_sample * 100,
    percentage_of_ids_per_sample_n_modif = n_identifications / n_ids_per_sample_n_modif * 100
  ) 
```

## Total Number of Identifications per Sample

```{r total_ids_plot}
#| fig-width: 8
#| fig-height: 20

# Plot total identifications per sample
tabular_nr_of_ids_per_sample %>%
  left_join(exploratory_results$experimental_design, by = "sample") %>%
  ggplot(aes(x = reorder(sample, n_ids_per_sample), 
             y = n_ids_per_sample,
             fill = condition)) +
  geom_col() +
  labs(
    title = "Total Number of Identifications per Sample",
    x = "Sample",
    y = "Number of identifications\n(N-term modif + peptide)",
    fill = "Condition"
  ) +
  coord_flip() +
  theme_minimal() + 
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )
```

## Identifications by N-terminal Modification and Preceding Amino Acid

```{r ids_by_modification_detailed}
#| fig-width: 8
#| fig-height: 20

# Plot identifications by modification type
tabular_nr_of_ids_per_aa_before %>% 
  left_join(exploratory_results$experimental_design, by = "sample") %>%
  ggplot(aes(x = reorder(sample, n_identifications),
             y = n_identifications,
             fill = p1_residue)) +
  facet_wrap(.~ nterm_modif, ncol = 2) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of IDs per N-terminal Modification and Preceding Amino Acid",
    x = "Sample",
    y = "Number of identifications\n(N-term modif + peptide)",
    fill = "Preceding residue"
  ) +
  theme_minimal() + 
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )
```

## Identifications by N-terminal Modification (Excluding R and K)

```{r ids_by_modification_no_rk}
#| fig-width: 14
#| fig-height: 20

# Plot identifications excluding R and K residues
tabular_nr_of_ids_per_aa_before %>%
  filter(p1_residue != "R", p1_residue != "K") %>%
  left_join(exploratory_results$experimental_design, by = "sample") %>%
  ggplot(aes(x = reorder(sample, n_identifications),
             y = n_identifications,
             fill = p1_residue)) +
  facet_wrap(.~ nterm_modif, ncol = 2) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Number of IDs per N-terminal Modification and Preceding Amino Acid",
    subtitle = "Excluding R and K residues",
    x = "Sample",
    y = "Number of identifications\n(N-term modif + peptide)",
    fill = "Preceding residue"
  ) +
  theme_minimal() + 
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )
```

## Proportion of Identifications by Preceding Amino Acid

```{r ids_proportion_aa}
#| fig-width: 15
#| fig-height: 20

# Plot proportions with R residue percentages
tabular_nr_of_ids_per_aa_before %>% 
  left_join(exploratory_results$experimental_design, by = "sample") %>%
  ggplot(aes(x = reorder(sample, percentage_of_ids_per_sample_n_modif),
             y = percentage_of_ids_per_sample_n_modif,
             fill = p1_residue)) +
  facet_wrap(.~ nterm_modif, ncol = 2) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Proportion of IDs per N-terminal Modification and Preceding Amino Acid",
    caption = "Numbers on the plot represent % of R residues at preceding amino acid in the sample",
    x = "Sample",
    y = "% of total identifications per sample\n(N-term modif + peptide)",
    fill = "Preceding residue"
  ) +
  # add a text label on top only of the R residues
  theme_minimal() + 
  theme(
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )
```

## Median Identifications by Processing Type

```{r}
# Calculate identifications by specificity and refined processing type (excluding specific peptides)

ids_by_specificity_refined <- exploratory_results$annotated_df_quant %>%
  filter(specificity %in% c("semi_Cterm", "semi_Nterm")) %>%
  dplyr::select(specificity, processing_type, nterm_modif, all_of(exploratory_results$experimental_design$sample)) %>%
  pivot_longer(cols = all_of(exploratory_results$experimental_design$sample),
               names_to = "sample",
               values_to = "value") %>%
  group_by(specificity, nterm_modif, processing_type, sample) %>%
  summarise(n_identifications = sum(!is.na(value)), .groups = 'drop') %>%
  left_join(
    exploratory_results$experimental_design, 
    by = "sample")

# Calculate median identifications by refined processing type
median_ids_refined_processing <- ids_by_specificity_refined %>%
  group_by(specificity, nterm_modif, processing_type, condition) %>%
  summarise(
    median_ids = median(n_identifications),
    mean_ids = mean(n_identifications),
    sd_ids = sd(n_identifications),
    n_samples = n(),
    .groups = 'drop'
  )

```

```{r median_ids_processing}
#| fig-width: 14
#| fig-height: 10

# Plot median identifications with facet_grid (horizontal bars)
plot_median_refined <- median_ids_refined_processing %>%
  ggplot(aes(x = median_ids, 
             y = condition,
             fill = condition)) +
  geom_col() +
  geom_errorbar(aes(xmin = median_ids - sd_ids, 
                    xmax = median_ids + sd_ids),
                width = 0.3, alpha = 0.7) +
  geom_text(aes(label = round(median_ids, 0)), 
            hjust = -0.1, vjust = -1, size = 3) +
  facet_grid(processing_type ~ nterm_modif + specificity, scales = "free_x") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Median Number of Identifications by Processing Type",
    subtitle = "Semi-specific peptides only, grouped by specificity and processing type",
    x = "Median number of identifications",
    y = "Condition",
    fill = "Condition"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(angle = 0, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(size = 9)
  )

print(plot_median_refined)

# Display summary table
knitr::kable(median_ids_refined_processing, 
             caption = "Summary Statistics: Identifications by Refined Processing Type and Condition",
             digits = 1)
```

# Proportional Abundances Analysis

```{r}
# extract peptide level data from summarized experiment object

mat_df_rawpur_nona <- as.data.frame(assay(exploratory_results$se_pept_imp_raw))

col_dat_df_rawpur_nona <- data.frame(colData(exploratory_results$se_pept_imp_raw))

row_dat_df_rawpur_nona <- data.frame(rowData(exploratory_results$se_pept_imp_raw))

# transform into long format
mat_df_rawpur_nona_long <- mat_df_rawpur_nona %>%
  tibble::rownames_to_column("nterm_modif_peptide") %>%
  tidyr::pivot_longer(cols = where(is.numeric), 
                      names_to = "sample", 
                      values_to  = "Intensity") %>%
  dplyr::left_join(., col_dat_df_rawpur_nona, 
                   by = "sample") %>%
  dplyr::left_join(., row_dat_df_rawpur_nona,
                   by = "nterm_modif_peptide") 
```

```{r}
# filter long data to keep only interesting products
mat_df_rawpur_nona_long_proteolytic <- mat_df_rawpur_nona_long %>%
  dplyr::filter(
    specificity %in% params$interesting_specificity, # exclude peptides with a preceding R or K
    nterm_modif %in% params$interesting_nterm_modif) %>% # keep only acetyl or unmodified peptides
  dplyr::mutate(
    condition = factor(condition, 
                         levels = condition_levels)
  )

pept_summ_rawpur_semi_1 <- mat_df_rawpur_nona_long_proteolytic %>% 
  group_by(sample, condition) %>%
  summarise(`Summed Abundances Semis` = sum(Intensity, na.rm = TRUE)) 

# get the summ of all peptides per sample/condition 
pept_summ_rawpur_all <- mat_df_rawpur_nona_long %>% 
  group_by(sample, condition) %>%
  summarise(`Summer Abundances Total` = sum(Intensity, na.rm = TRUE)) 

# merge the two data frames to get the normalized abundances of proteolytic products
pept_summ_rawpur_semi_3 <- pept_summ_rawpur_semi_1 %>%
  dplyr::left_join(., pept_summ_rawpur_all, 
                   by = c("sample", "condition")) %>%
  mutate(`% of Semi/Total Abundances` = `Summed Abundances Semis`/`Summer Abundances Total` * 100 ) %>%
  mutate(
    condition = factor(condition, 
                       levels = condition_levels)
    ) %>%
  left_join(., 
            experimental_design %>% 
              dplyr::select(sample, bio_replicate), 
            by = "sample")
```

```{r proportion_plots}
#| fig-width: 12
#| fig-height: 5

# First plot: points and lines
plot_lines <- ggplot(pept_summ_rawpur_semi_3, 
  aes(x = condition, 
      y = `% of Semi/Total Abundances`,
      color = condition)) +
  geom_point(aes(group = bio_replicate), 
             alpha = 0.7, 
             size = 2) +
  geom_line(aes(group = bio_replicate), 
            alpha = 0.5, 
            color = "gray50") +
  labs(x = "Condition",
       y = "% Abundance of Proteolytic products",
       title = "Proportional abundances of proteolytic products - Connected by replicate") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Second plot: boxplot with jittered points
plot_boxplot <- ggplot(pept_summ_rawpur_semi_3, 
  aes(x = condition, 
      y = `% of Semi/Total Abundances`,
      fill = condition)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, 
              size = 1,
              width = 0.2) +
  labs(x = "Condition",
       y = "% Abundance of Proteolytic products",
       title = "Proportional abundances of proteolytic products - Distribution") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(plot_lines)
print(plot_boxplot)
```

# Differential Abundance Analysis Results

## Prepare volcano plot data

```{r prepare_volcano_data}
# Function to prepare volcano data
prepare_volcano_data <- function(data, fc_thresh = fc_threshold, p_thresh = pval_threshold, contrast_levels = contrast_levels) {
  df <- data %>%
    mutate(
      `-log10_adj.P.Val` = -log10(adj.P.Val),
      regulation = case_when(
        logFC > fc_thresh & adj.P.Val < p_thresh ~ "upregulated",
        logFC < -fc_thresh & adj.P.Val < p_thresh ~ "downregulated",
        TRUE ~ "non-changing"
      ),
      contrast = factor(
        contrast,
        levels = contrast_levels
      )
    )

    # evaluate proper definition of contrast levels
    if (any(is.na(df$contrast))) {

      stop("At least some contrast levels in the data are not properly defined in contrast_levels parameter. Please check.")
    
    }
  
  return(df)
}

# Prepare data for all analyses
volcano_data_before <- prepare_volcano_data(
  limma_results$by_dataset$non_protnorm$results, 
  contrast_levels = contrast_levels)

volcano_data_after <- prepare_volcano_data(
  limma_results$by_dataset$protnorm_with_lost$results, 
  contrast_levels = contrast_levels)

volcano_data_protein <- prepare_volcano_data(
  limma_results$by_dataset$protein_abund_norm$results, 
  contrast_levels = contrast_levels)
```

## Volcano Plots

```{r volcano_plots}
#| fig-width: 15
#| fig-height: 5

# Function to create volcano plot
create_volcano_plot <- function(data, title_text, subtitle_text) {
  # Count significant features
  counts <- data %>%
    dplyr::filter(regulation != "non-changing") %>%
    group_by(contrast) %>%
    summarise(count = n(), .groups = 'drop')
  
  ggplot(data, aes(x = logFC, y = `-log10_adj.P.Val`, color = regulation)) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = -log10(pval_threshold), 
               linetype = "dashed", color = "red") +
    scale_color_manual(values = c("upregulated" = "red", 
                                  "downregulated" = "blue", 
                                  "non-changing" = "grey")) +
    facet_wrap(~ contrast, ncol = 3) +
    theme_minimal() +
    labs(title = title_text,
         subtitle = subtitle_text,
         x = "Log Fold Change", 
         y = "-log10(adjusted P-value)", 
         color = "Regulation") +
    theme(strip.text.x = element_text(size = 10),
          axis.text = element_text(size = 8),
          axis.title = element_text(size = 10),
          legend.position = "bottom") +
    geom_text(data = counts, aes(x = Inf, y = Inf, label = count),
              hjust = 1.1, vjust = 1.5, size = 4, color = "black", inherit.aes = FALSE)
}

# Create volcano plots
volcano_before <- create_volcano_plot(
  volcano_data_before,
  "Differential Abundance: Before Protein Normalization",
  "Neo-termini analysis"
)

volcano_after <- create_volcano_plot(
  volcano_data_after,
  "Differential Abundance: After Protein Normalization", 
  "Neo-termini analysis"
)

volcano_protein <- create_volcano_plot(
  volcano_data_protein,
  "Differential Abundance: Protein Level",
  "Based on fully specific peptides"
)

print(volcano_before)
print(volcano_after)
print(volcano_protein)
```

## Summary Statistics

```{r summary_stats}
# Create summary of differential features
create_summary <- function(data, analysis_name) {
  data %>%
    filter(regulation != "non-changing") %>%
    group_by(contrast, regulation) %>%
    summarise(count = n(), .groups = 'drop') %>%
    mutate(analysis = analysis_name)
}

summary_stats <- bind_rows(
  create_summary(volcano_data_before, "Before Protein Norm"),
  create_summary(volcano_data_after, "After Protein Norm"),
  create_summary(volcano_data_protein, "Protein Level")
)

# Plot summary
ggplot(summary_stats, aes(x = contrast, y = count, fill = regulation)) +
  geom_col(position = position_dodge(width = 0.8)) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.5, size = 3) +
  facet_wrap(~ analysis, scales = "free_y") +
  scale_fill_manual(values = c("upregulated" = "red", "downregulated" = "blue")) +
  labs(title = "Summary of Differentially Abundant Features",
       x = "Contrast", y = "Number of Features", fill = "Regulation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

# Principal Component Analysis

```{r pca_analysis}
#| fig-width: 14
#| fig-height: 14

# Function to perform PCA and create plot
perform_pca_plot <- function(abundance_matrix, title_text, sample_data = sample_annotation) {
  # Transpose matrix for PCA
  t_matrix <- t(abundance_matrix)
  
  # Perform PCA
  pca_res <- pca(t_matrix, ncomp = 2)
  
  # Prepare data for plotting
  pca_df <- pca_res$x[, 1:2] %>%
    as.data.frame() %>%
    rownames_to_column(var = "sample") %>%
    left_join(sample_data, by = "sample")
  
  # Get variance explained
  var_exp_pc1 <- pca_res$prop_expl_var$X[1]
  var_exp_pc2 <- pca_res$prop_expl_var$X[2]
  
  # Create plot
  ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
    geom_point(size = 3.5) +
    stat_ellipse(level = 0.50) +
    labs(title = title_text,
         x = paste("PC1 - Variance explained:", round(var_exp_pc1, 3) * 100, "%"),
         y = paste("PC2 - Variance explained:", round(var_exp_pc2, 3) * 100, "%"),
         color = "Condition") +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# Create PCA plots
pca_before <- perform_pca_plot(
  assay(exploratory_results$se_pept_non_protnorm),
  "PCA: Neo-termini Before Protein Normalization",
  sample_data = experimental_design
)

pca_after <- perform_pca_plot(
  assay(exploratory_results$se_pept_protnorm_with_lost),
  "PCA: Neo-termini After Protein Normalization",
  sample_data = experimental_design
)

pca_protein <- perform_pca_plot(
  assay(exploratory_results$se_prot_norm),
  "PCA: Protein Abundance Data",
  sample_data = experimental_design
)

# Arrange plots
ggarrange(
  pca_before, 
  pca_after, 
  pca_protein, 
  ncol = 2, 
  nrow = 2, 
  common.legend = TRUE, 
  legend = "bottom")
```

# Principal Component Analysis - Comprehensive Analysis

```{r pca_comprehensive}
#| fig-width: 16
#| fig-height: 22

# Create all PCA plots
pca_neo_before <- perform_pca_plot(
  exploratory_results$se_pept_non_protnorm[
  # filter row data to only neo-termini
  rowData(exploratory_results$se_pept_non_protnorm)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
  rowData(exploratory_results$se_pept_non_protnorm)$specificity %in% exploratory_results$parameters$interesting_specificity
    ] %>%
    assay(),
  "All Neo-termini (Before Protein Norm)",
  sample_data = experimental_design
)

pca_neo_after <- perform_pca_plot(
  exploratory_results$se_pept_protnorm_with_lost[
  # filter row data to only neo-termini
  rowData(exploratory_results$se_pept_protnorm_with_lost)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
  rowData(exploratory_results$se_pept_protnorm_with_lost)$specificity %in% exploratory_results$parameters$interesting_specificity
    ] %>%
    assay(),
  "All Neo-termini (After Protein Norm)",
  sample_data = experimental_design
)

pca_semi_c_before <- perform_pca_plot(
  exploratory_results$se_pept_non_protnorm[
    # filter row data to only neo-termini
    rowData(exploratory_results$se_pept_non_protnorm)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
    rowData(exploratory_results$se_pept_non_protnorm)$specificity %in% "semi_Cterm"
    ] %>%
    assay(),
  "Semi C-terminal (Before Protein Norm)",
  sample_data = experimental_design
)

pca_semi_c_after <- perform_pca_plot(
  exploratory_results$se_pept_protnorm_with_lost[
    # filter row data to only neo-termini
    rowData(exploratory_results$se_pept_protnorm_with_lost)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
    rowData(exploratory_results$se_pept_protnorm_with_lost)$specificity %in% "semi_Cterm"
    ] %>%
    assay(),
  "Semi C-terminal (After Protein Norm)",
  sample_data = experimental_design
)

pca_semi_n_before <- perform_pca_plot(
  exploratory_results$se_pept_non_protnorm[
    # filter row data to only neo-termini
    rowData(exploratory_results$se_pept_non_protnorm)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
    rowData(exploratory_results$se_pept_non_protnorm)$specificity %in% "semi_Nterm"
    ] %>%
    assay(),
  "Semi N-terminal (Before Protein Norm)",
  sample_data = experimental_design
)

pca_semi_n_after <- perform_pca_plot(
  exploratory_results$se_pept_protnorm_with_lost[
    # filter row data to only neo-termini
    rowData(exploratory_results$se_pept_protnorm_with_lost)$nterm_modif %in% exploratory_results$parameters$interesting_nterm_modif & 
    rowData(exploratory_results$se_pept_protnorm_with_lost)$specificity %in% "semi_Nterm"
    ] %>%
    assay(),
  "Semi N-terminal (After Protein Norm)",
  sample_data = experimental_design
)

pca_protein <- perform_pca_plot(
  exploratory_results$se_prot_norm %>%
    assay(),
  "Protein Abundance Data",
  sample_data = experimental_design
)

# Arrange all plots
ggarrange(
  pca_neo_before, pca_neo_after,
  pca_semi_c_before, pca_semi_c_after,
  pca_semi_n_before, pca_semi_n_after,
  pca_protein, NULL,
  ncol = 2, nrow = 4, 
  common.legend = TRUE, legend = "bottom"
)
```

# Intersection Analysis: Before vs After Protein Normalization

```{r intersection_analysis}
#| fig-width: 10
#| fig-height: 4

if (!is.null(all_results) && all(c("dataset", "contrast", "nterm_modif_peptide", "logFC", "adj.P.Val") %in% names(all_results))) {

  contrasts <- unique(all_results$contrast)

  make_sets <- function(contrast_name, direction = c("up", "down")) {

    direction <- match.arg(direction)
    df <- all_results %>% dplyr::filter(contrast == contrast_name)

    if (direction == "up") {
      before <- df %>% dplyr::filter(
        dataset == "non_protnorm",
        adj.P.Val < pval_threshold,
        logFC > log2(fc_threshold)
      )
      after <- df %>% dplyr::filter(
        dataset == "protnorm_with_lost",
        adj.P.Val < pval_threshold,
        logFC > log2(fc_threshold)
      )
    } else {
      before <- df %>% dplyr::filter(
        dataset == "non_protnorm",
        adj.P.Val < pval_threshold,
        logFC < -log2(fc_threshold)
      )
      after <- df %>% dplyr::filter(
        dataset == "protnorm_with_lost",
        adj.P.Val < pval_threshold,
        logFC < -log2(fc_threshold)
      )
    }

    list(
      Before = unique(before$nterm_modif_peptide),
      After = unique(after$nterm_modif_peptide)
    )
  }

  for (cn in contrasts) {
    # Upregulated
    sets_up <- make_sets(cn, "up")
    comb_up <- ComplexHeatmap::make_comb_mat(sets_up)
    ta_up <- ComplexHeatmap::upset_top_annotation(comb_up, add_numbers = TRUE)
    ra_up <- ComplexHeatmap::upset_right_annotation(comb_up, add_numbers = TRUE)
    print(ComplexHeatmap::UpSet(
      comb_up,
      set_order = c("Before", "After"),
      column_title = paste0("Upregulated - ", cn),
      top_annotation = ta_up,
      right_annotation = ra_up
    ))

    # Downregulated
    sets_down <- make_sets(cn, "down")
    comb_down <- ComplexHeatmap::make_comb_mat(sets_down)
    ta_down <- ComplexHeatmap::upset_top_annotation(comb_down, add_numbers = TRUE)
    ra_down <- ComplexHeatmap::upset_right_annotation(comb_down, add_numbers = TRUE)
    print(ComplexHeatmap::UpSet(
      comb_down,
      set_order = c("Before", "After"),
      column_title = paste0("Downregulated - ", cn),
      top_annotation = ta_down,
      right_annotation = ra_down
    ))
  }
} else {
  message("Required columns for intersection analysis not found; skipping.")
}
```

# Cleavage Pattern Analysis (requires user input, project-specific)

## Prepare gene-ontology annotation (for determining intra- vs extracellular)

```{r}
# prepare gene-ontology grouping 

go_annotation_df <- as.data.frame(go_groups_df$protnorm_with_lost@result) %>%
  dplyr::rename(
    go_id = ID,
    go_description = Description,
    gene = geneID
  ) %>%
  dplyr::select(
    go_id,
    go_description,
    gene
  ) %>%
  distinct() %>%
  separate_rows(gene, sep = "/") %>%
  dplyr::filter(
    !is.na(gene),
    gene != ""
  )

# merge GO annotation with volcano data

volcano_data_after_w_go <- volcano_data_after %>%
  left_join(
    protein_annotation %>%
      dplyr::select(protein, gene),
    by = c("protein" = "protein")
  ) %>%
  left_join(
    go_annotation_df,
    by = c("gene" = "gene")
  ) %>% 
  # collapse the go_id and go_description columns to have all terms per protein in one row
  group_by(
    across(-c(go_id, go_description))) %>%
  summarise(
    go_id = paste(unique(go_id), collapse = "; "),
    go_description = paste(unique(go_description), collapse = "; "),
    .groups = "drop"
  ) %>%
  mutate(
    extracellular_or_cytoplasm = case_when(
      str_detect(go_description, "extracellular") ~ "extracellular",
      str_detect(go_description, "cytoplasm|intracellular") ~ "intracellular|cytoplasm",
      TRUE ~ "other"
    )
  )
```

## Prepare Data for Cleavage Analysis

```{r cleavage_data_prep}
# Function to extract cleavage sequences for each contrast
get_upreg_cleavage_sequences <- function(
  contrast_name,
  regulation_type = "upregulated",
  nterm_modif = "all",
  processing_type = "not_canonical"
  ) {

  build_sets_for_modif <- function(modif_value) {

    base_df <- volcano_data_after_w_go %>%
      filter(
        contrast == contrast_name,
        regulation == regulation_type,
        processing_type == processing_type,
        nterm_modif == modif_value,
        p1_position > 2 # avoid N-terminal residues
      ) %>%
      mutate(cleavage_sequence_x = paste0(x_res_before, x_res_after))

    extracellular <- base_df %>%
      filter(str_detect(extracellular_or_cytoplasm, "extracellular"))
    intracellular <- base_df %>%
      filter(str_detect(extracellular_or_cytoplasm, "intracellular|cytoplasm"))

    list(
      contrast = contrast_name,
      nterm_modif = modif_value,
      c_term_int = intracellular %>%
        filter(specificity == "semi_Cterm") %>%
        pull(cleavage_sequence_x),
      c_term_ext = extracellular %>%
        filter(specificity == "semi_Cterm") %>%
        pull(cleavage_sequence_x),
      n_term_int = intracellular %>%
        filter(specificity == "semi_Nterm") %>%
        pull(cleavage_sequence_x),
      n_term_ext = extracellular %>%
        filter(specificity == "semi_Nterm") %>%
        pull(cleavage_sequence_x)
    )
  }

  if (nterm_modif == "all") {

    modifs <- volcano_data_after_w_go %>%
      dplyr::pull(nterm_modif) %>%
      unique() %>%
      stats::na.omit() %>%
      as.character()

    modifs <- sort(modifs)

    return(
      purrr::map(modifs, build_sets_for_modif) %>%
        purrr::set_names(modifs)
    )
  }

  build_sets_for_modif(nterm_modif)
}

# Loop over contrasts defined in the first chunk
cleavage_sequences_by_contrast <- purrr::map(
  contrast_levels,
  get_upreg_cleavage_sequences,
  regulation_type = "upregulated",
  nterm_modif = "all",
  processing_type = "not_canonical"
) %>%
  purrr::set_names(contrast_levels)
```

## Cleavage Pattern Heatmaps

```{r cleavage_heatmaps}
#| fig-width: 6
#| fig-height: 4
#| cache: false

# Helper to plot heatmaps for a given sequence list
plot_cleavage_heatmaps <- function(seq_list, contrast_name, modif_name, p_number = 10) {

  # Build count matrices safely (empty vectors return NULL)
  mk_counts <- function(x) {
    if (length(x) == 0) return(NULL)
    cleavage_area_matrix(x, p_number = p_number)$amino_acid_count
  }

  counts <- list(
    c_int = mk_counts(seq_list$c_term_int),
    c_ext = mk_counts(seq_list$c_term_ext),
    n_int = mk_counts(seq_list$n_term_int),
    n_ext = mk_counts(seq_list$n_term_ext)
  )

  op <- par(mfrow = c(2, 2))
  on.exit(par(op), add = TRUE)

  plotted_any <- FALSE

  if (!is.null(counts$c_int)) {
    p <- pheatmap(
      counts$c_int,
      cluster_rows = FALSE, cluster_cols = FALSE,
      main = sprintf("%s | %s\nIntracellular C-terminal (n=%s)", contrast_name, modif_name, length(seq_list$c_term_int)),
      color = colorRampPalette(brewer.pal(n = 9, name = "Reds"))(100)
    )
    print(p)
    plotted_any <- TRUE
  }
  if (!is.null(counts$c_ext)) {
    p <- pheatmap(
      counts$c_ext,
      cluster_rows = FALSE, cluster_cols = FALSE,
      main = sprintf("%s | %s\nExtracellular C-terminal (n=%s)", contrast_name, modif_name, length(seq_list$c_term_ext)),
      color = colorRampPalette(brewer.pal(n = 9, name = "Blues"))(100)
    )
    print(p)
    plotted_any <- TRUE
  }
  if (!is.null(counts$n_int)) {
    p <- pheatmap(
      counts$n_int,
      cluster_rows = FALSE, cluster_cols = FALSE,
      main = sprintf("%s | %s\nIntracellular N-terminal (n=%s)", contrast_name, modif_name, length(seq_list$n_term_int)),
      color = colorRampPalette(brewer.pal(n = 9, name = "Greens"))(100)
    )
    print(p)
    plotted_any <- TRUE
  }
  if (!is.null(counts$n_ext)) {
    p <- pheatmap(
      counts$n_ext,
      cluster_rows = FALSE, cluster_cols = FALSE,
      main = sprintf("%s | %s\nExtracellular N-terminal (n=%s)", contrast_name, modif_name, length(seq_list$n_term_ext)),
      color = colorRampPalette(brewer.pal(n = 9, name = "Purples"))(100)
    )
    print(p)
    plotted_any <- TRUE
  }

  if (!plotted_any) {
    message("No cleavage heatmaps to plot for ", contrast_name, " | ", modif_name)
  }
}

# Iterate over contrasts and nterm_modif elements
purrr::iwalk(
  cleavage_sequences_by_contrast,
  function(modif_list, contrast_nm) {
    purrr::iwalk(
      modif_list,
      function(seq_list, modif_nm) {
        plot_cleavage_heatmaps(seq_list, contrast_name = contrast_nm, modif_name = modif_nm)
      }
    )
  }
)
```

## Sequence Logos

```{r sequence_logos}
#| fig-width: 8
#| fig-height: 5

if (!isTRUE(run_sequence_logos)) {
  message("Sequence logos skipped (run_sequence_logos = FALSE).")
} else {
  # Prepare proteome for dagLogo analysis
  proteome <- prepareProteome(fasta = fasta_location, species = "Homo sapiens")

  # Function to generate dagLogo if sufficient sequences
  generate_daglogo <- function(sequences, title_text, min_seq = 5) {
    if (length(sequences) >= min_seq) {
      set.seed(404)

      seqs_4ice <- formatSequence(seq = sequences, proteome = proteome)

      seqs_4ice_ztest <- buildBackgroundModel(
        seqs_4ice,
        model = "any",
        background = "wholeProteome",
        proteome = proteome,
        testType = "ztest"
      )

      seqs_4ice_ztest_dau <- testDAU(seqs_4ice, dagBackground = seqs_4ice_ztest)

      dagLogo(
        seqs_4ice_ztest_dau,
        labels = c(
          "P10", "P9", "P8", "P7", "P6", "P5", "P4", "P3", "P2", "P1",
          "P1'", "P2'", "P3'", "P4'", "P5'", "P6'", "P7'", "P8'", "P9'", "P10'"
        )
      )
    } else {
      plot.new()
      text(
        0.5,
        0.5,
        paste0(title_text, "\nInsufficient sequences (n=", length(sequences), ")"),
        cex = 1.2,
        adj = 0.5
      )
    }
  }

  # Helper to choose contrasts/modifs to plot
  contrast_keys <- names(cleavage_sequences_by_contrast)
  if (!is.null(sequence_logo_contrasts)) {
    contrast_keys <- intersect(contrast_keys, sequence_logo_contrasts)
  }

  purrr::iwalk(
    cleavage_sequences_by_contrast[contrast_keys],
    function(modif_list, contrast_nm) {
      modif_keys <- names(modif_list)
      if (!is.null(sequence_logo_nterm_modifs)) {
        modif_keys <- intersect(modif_keys, sequence_logo_nterm_modifs)
      }

      purrr::iwalk(
        modif_list[modif_keys],
        function(seq_list, modif_nm) {
          title_prefix <- paste0(contrast_nm, " | ", modif_nm)

          generate_daglogo(
            seq_list$c_term_int,
            paste0(title_prefix, " — Intracellular C-terminal")
          )
          generate_daglogo(
            seq_list$c_term_ext,
            paste0(title_prefix, " — Extracellular C-terminal")
          )
          generate_daglogo(
            seq_list$n_term_int,
            paste0(title_prefix, " — Intracellular N-terminal")
          )
          generate_daglogo(
            seq_list$n_term_ext,
            paste0(title_prefix, " — Extracellular N-terminal")
          )
        }
      )
    }
  )
}
```

# Histogram of Cleavage Positions

```{r cleavage_histograms}
#| fig-width: 8
#| fig-height: 7

if (all(c("p1_prime_position", "protein_length", "dataset", "contrast") %in% names(all_results))) {

  build_cleavage_histogram <- function(df, contrast_name, title_text, facet_col) {
    if (nrow(df) == 0) return(NULL)

    ggplot(df, aes(x = normalized_cleavage_position)) +
      geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
      facet_grid(nterm_modif + .data[[facet_col]] ~ m_in_p1_or_p1_prime) +
      labs(
        title = title_text,
        subtitle = contrast_name,
        x = "Normalized Cleavage Site Position (%)",
        y = "Count"
      ) +
      theme_minimal() +
      theme(
        strip.text = element_text(size = 10),
        plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 12)
      )
  }

  # Container for plots
  cleavage_histogram_plots <- list()

  mode_specs <- list(
    updown = list(
      title = "Histogram of Cleavage Site Positions\nUpregulated and Downregulated Non-canonical Features",
      df_fn = function(cn) {
        volcano_data_after_w_go %>%
          dplyr::filter(
            contrast == cn,
            regulation != "not_significant",
            processing_type == "not_canonical",
            !is.na(p1_prime_position),
            !is.na(protein_length),
            protein_length > 0
          ) %>%
          dplyr::mutate(
            normalized_cleavage_position = p1_prime_position / protein_length * 100,
            m_in_p1_or_p1_prime = case_when(
              str_detect(p1_residue, "M") | str_detect(p1_prime_residue, "M") ~ "M in P1 or P1'",
              TRUE ~ "no M in P1 or P1'"
            )
          )
      },
      facet_col = "extracellular_or_cytoplasm"
    ),
    upregulated = list(
      title = "Histogram of Cleavage Site Positions\nUpregulated Non-canonical Features",
      df_fn = function(cn) {
        volcano_data_after_w_go %>%
          dplyr::filter(
            dataset == "protnorm_with_lost",
            contrast == cn,
            regulation == "upregulated",
            processing_type == "not_canonical",
            !is.na(p1_prime_position),
            !is.na(protein_length),
            protein_length > 0
          ) %>%
          dplyr::mutate(
            normalized_cleavage_position = p1_prime_position / protein_length * 100,
            m_in_p1_or_p1_prime = case_when(
              str_detect(p1_residue, "M") | str_detect(p1_prime_residue, "M") ~ "M in P1 or P1'",
              TRUE ~ "no M in P1 or P1'"
            )
          )
      },
      facet_col = "extracellular_or_cytoplasm"
    ),
    downregulated = list(
      title = "Histogram of Cleavage Site Positions\nDownregulated Non-canonical Features",
      df_fn = function(cn) {
        volcano_data_after_w_go %>%
          dplyr::filter(
            dataset == "protnorm_with_lost",
            contrast == cn,
            regulation == "downregulated",
            processing_type == "not_canonical",
            !is.na(p1_prime_position),
            !is.na(protein_length),
            protein_length > 0
          ) %>%
          dplyr::mutate(
            normalized_cleavage_position = p1_prime_position / protein_length * 100,
            m_in_p1_or_p1_prime = case_when(
              str_detect(p1_residue, "M") | str_detect(p1_prime_residue, "M") ~ "M in P1 or P1'",
              TRUE ~ "no M in P1 or P1'"
            )
          )
      },
      facet_col = "extracellular_or_cytoplasm"
    ),
    all_features = list(
      title = "Histogram of Cleavage Site Positions\nAll Non-canonical Features (incl. non-DA)",
      df_fn = function(cn) {
        volcano_data_after_w_go %>%
          dplyr::filter(
            dataset == "protnorm_with_lost",
            contrast == cn,
            processing_type == "not_canonical",
            !is.na(p1_prime_position),
            !is.na(protein_length),
            protein_length > 0
          ) %>%
          dplyr::mutate(
            normalized_cleavage_position = p1_prime_position / protein_length * 100,
            m_in_p1_or_p1_prime = case_when(
              str_detect(p1_residue, "M") | str_detect(p1_prime_residue, "M") ~ "M in P1 or P1'",
              TRUE ~ "no M in P1 or P1'"
            )
          )
      },
      facet_col = "extracellular_or_cytoplasm"
    )
  )

  for (contrast in unique(volcano_data_after$contrast)) {

    message("Building cleavage histograms for contrast: ", contrast)

    for (mode_name in names(mode_specs)) {
      spec <- mode_specs[[mode_name]]
      df_plot <- spec$df_fn(contrast)

      facet_col <- spec$facet_col
      if (!facet_col %in% names(df_plot)) {
        facet_col <- if ("intra_or_extra" %in% names(df_plot)) "intra_or_extra" else "nterm_modif"
      }

      p <- build_cleavage_histogram(df_plot, contrast, spec$title, facet_col)

      if (!is.null(p)) {
        plot_key <- paste(contrast, mode_name, sep = "__")
        cleavage_histogram_plots[[plot_key]] <- p
      }
    }
  }

} else {
  message("Required columns for cleavage position histogram not found; skipping.")
}
```

```{r}
#| fig-width: 8
#| fig-height: 7

if (length(cleavage_histogram_plots) > 0) {
  purrr::iwalk(
    cleavage_histogram_plots,
    function(plot_obj, plot_name) {
      print(plot_obj)
    }
  )
} else {
  message("No cleavage histogram plots were generated.")
}
```

# Save Plots and Summary

```{r save_results}
# Save summary statistics
write_tsv(
  summary_stats,
  file.path(results_dir, paste0(pre_fix, "_summary_statistics.tsv"))
)

message("Visualization analysis complete. Results saved.")
```

# Session Info

```{r session_info}
sessionInfo()
```

```{r}
str(go_groups_df$protnorm_with_lost@result)
```